---
title: "compare vga iterations"
author: "DongyueXie"
date: "2023-01-02"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

The vga step is an optimization problem. 1 iteration can already increase the elbo. Running till convergence will increase more elbo but is that necessary? does this benefit the overall fit in the long run?

I compare VGA 1 iteration vs convergence

```{r}
res1000 = readRDS('/project2/mstephens/dongyue/poisson_mf/compare_vga_iterations/pbmc_3cells_vga1000.rds')
res1 = readRDS('/project2/mstephens/dongyue/poisson_mf/compare_vga_iterations/pbmc_3cells_vga1.rds')
```

## compare elbo

```{r}
plot(res1$fit$elbo_trace,type='l',col=2,lwd=2)
lines(res1000$fit$elbo_trace,col=4,lwd=2)
legend('bottomright',c('vga 1iter','vga converge'),lty=c(1,1),col=c(2,4))
res1$fit$elbo
res1000$fit$elbo
```

Not much difference. Running 1 iteration actually results in a slightly larger elbo.

## compare run time

```{r}
res1$fit$run_time
res1000$fit$run_time
```

```{r}
lapply(res1$fit$run_time_break_down,mean)$run_time_vga
lapply(res1000$fit$run_time_break_down,mean)$run_time_vga
```


1 iteration is much faster.

## fitted sturcture

```{r}
res1$fit$fit_flash$n.factors
res1000$fit$fit_flash$n.factors
```

```{r}
library(fastTopics)
data(pbmc_facs)

### use three cells
cells = pbmc_facs$samples$subpop%in%c('B cell', 'NK cell','CD34+')
cell_names =pbmc_facs$samples$subpop[cells]
```

```{r}
source('code/poisson_STM/plot_factors.R')
```

```{r}
plot.factors(res1$fit$fit_flash,cell_names,title='vga 1 iter')
plot.factors(res1000$fit$fit_flash,cell_names,title='vga converge')
```

The last 3 factors are slightly different.

## Other concerns

The 1 iteration version requires storing the full dense matrix V which can take huge memory space when the dimension is large. For example for a $10^5\times 10^3$ dense matrix it takes > 16GB RAM. 

The converged version can exploit the explicit relationship between optimal M and V so storing V is not required.
