---
title: "ebpmf init using liklihood expansion"
author: "DongyueXie"
date: "2023-07-22"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction


Consider a Poisson distribution, $y\sim Poisson(s\exp(\mu))$, the log likelihood of $\mu$ is 
\[l(\mu) = y\mu - s\exp\mu\]

A second order Taylor series expansion of $\mu$ around 0 is 
\[\tilde l(\mu) = -s + (y-s)\mu - \frac{s}{2}\mu^2 = -\frac{s}{2}(\mu-\frac{y-s}{s})^2.\]

So the ``pseudo-data'' are $(y-s)/s$.

Remark:

If in another model $y\sim Poisson(\exp\mu)$, and we want to perform a 2nd order Taylor series expansion of $\mu$ around $\alpha$, then the ``pseudo-data'' are $\alpha + \frac{y-\exp(\alpha)}{\exp(\alpha)}$.

We can fit an EBMF model to the pseudo-data as initialization.

```{r}
library(Matrix)
datax = readRDS('data/sla_2000.rds')
dim(datax$data)
sum(datax$data==0)/prod(dim(datax$data))
datax$data = Matrix(datax$data,sparse = TRUE)
```

```{r}
l0 = log(rowMeans(datax$data))
f0 = log(colSums(datax$data)/sum(exp(l0)))
alpha = l0%*%t(rep(1,ncol(datax$data))) + rep(1,nrow(datax$data))%*%t(f0)
Y_tilde = as.matrix((datax$data - exp(alpha))/exp(alpha))
```


```{r}
library(flashier)
fit = flash(as.matrix(Y_tilde),ebnm=ebnm::ebnm_point_exponential,greedy.Kmax = 6,var.type = 2,backfit = T,S = sqrt(1/exp(alpha)))
```



```{r}
colnames(datax$data)[order(fit$F.pm[,1],decreasing = T)[1:10]]
colnames(datax$data)[order(fit$F.pm[,2],decreasing = T)[1:10]]
colnames(datax$data)[order(fit$F.pm[,3],decreasing = T)[1:10]]
colnames(datax$data)[order(fit$F.pm[,4],decreasing = T)[1:10]]
colnames(datax$data)[order(fit$F.pm[,5],decreasing = T)[1:10]]
colnames(datax$data)[order(fit$F.pm[,6],decreasing = T)[1:10]]
```




```{r}
fit1 = flash(as.matrix(Y_tilde)/sqrt(1/exp(alpha)),ebnm=ebnm::ebnm_point_exponential,greedy.Kmax = 6,var.type = 2,backfit = T)
```


```{r}
colnames(datax$data)[order(fit1$F.pm[,1],decreasing = T)[1:10]]
colnames(datax$data)[order(fit1$F.pm[,2],decreasing = T)[1:10]]
colnames(datax$data)[order(fit1$F.pm[,3],decreasing = T)[1:10]]
colnames(datax$data)[order(fit1$F.pm[,4],decreasing = T)[1:10]]
colnames(datax$data)[order(fit1$F.pm[,5],decreasing = T)[1:10]]
colnames(datax$data)[order(fit1$F.pm[,6],decreasing = T)[1:10]]
```

## summary


Does not work well. exp(alpha) can be very small. so pesudo-data can be very large.

