---
title: "Fit ebpmf on statisticalpaper abstracts (SLA) data"
author: "DongyueXie"
date: "2023-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Dataset is from \url{https://github.com/ZhengTracyKe/TopicSCORE/tree/main/Statistical%20Literature%20Abstracts%20data}



```{r}
library(flashier)
data <- read.table('data/abstract.txt')
data <- as.matrix(data)
vocab <- read.table('data/abstract.vocab.txt',colClasses = "character")[[1]]

# D_count is word by document matrix
D_count <- matrix(0,max(data[,2]),max(data[,1]))
for (t in 1:dim(data)[1]){
  D_count[data[t,2], data[t,1]] <-data[t,3]
}

p <- dim(D_count)[1]
n <- dim(D_count)[2]

# Set some thresholds
w_num <- 3000        #number of words to keep
d_percent <- 0.6     #percentage of docs to keep
Mquantile <- 1     #Truncate quantile of M

#Only keep d_percent% longest documents
doc_count <- colSums(D_count)
doc_keep <- which(rank(-doc_count, ties.method = 'first')<=round(d_percent*n))
D_count <- D_count[,doc_keep]

#Only keep top w_num most frequent words
word_count <- rowSums(D_count)
word_keep <- which(rank(-word_count, ties.method = 'first')<=w_num)
D_count <- D_count[word_keep,]
vocab <- vocab[word_keep]

dim(D_count)
hist(colSums(D_count),breaks = 100)

## run ebpmf
library(ebpmf)
fit = ebpmf_log(t(D_count),flash_control=list(Kmax=10,
                                               ebnm.fn=c(ebnm::ebnm_point_exponential,ebnm::ebnm_point_exponential),
                                               loadings_sign=1,
                                               factors_sign = 1),
                var_type = 'by_col',
                init_control = list(init_tol=1e-4,single_gene_expmix=TRUE,deal_with_no_init_factor='...'),
                sigma2_control = list(return_sigma2_trace=T),
                general_control = list(maxiter=1,conv_tol=1e-5,save_init_val=TRUE,save_latent_M=T))
```

## Why no factor is added?

We see in the ebpmf fit that at the initialization, the model fails to find new factors, other than the intercepts. 

We first extract the flash object, and reproduce the results.

```{r}
hist(fit$init_val$sigma2_init,breaks = 100)
range(fit$init_val$M_init)
# re-produce that flash could not add any more dimensions.
fit.flash = fit$fit_flash$flash.fit
fit.flash  = flash.add.greedy(fit.flash, Kmax = 10,ebnm.fn = ebnm::ebnm_point_exponential,verbose = 1)
```

How about just supply latent data to flash, and do not provide variances?

```{r}
fit_flash = flash.init(fit$init_val$M_init,S=NULL,var.type = 2)
l0 = log(rowMeans(t(D_count)))
n = ncol(D_count)
p = nrow(D_count)
ebnm.fixed.l0 = function(x,s,g_init,fix_g,output){
    return(list(posterior=list(mean=l0,second_moment = l0^2),
                fitted_g = NULL,
                log_likelihood=sum(dnorm(x,l0,s,log=T))))
  }
fit_flash = flash.init.factors(fit_flash,list(cbind(l0), cbind(rep(1,p))),ebnm.fn = ebnm.fixed.l0) %>%
    flash.fix.factors(kset = 1, mode = 2)
f0 = log(cbind(colSums(t(D_count))/sum(exp(l0))))
  fit_flash = flash.init.factors(fit_flash,list(cbind(rep(1,n)), f0),ebnm.fn = ebnm::ebnm_normal) %>%
    flash.fix.factors(kset = 2, mode = 1)

fit_flash = flash.add.greedy(fit_flash, Kmax = 10,ebnm.fn = ebnm::ebnm_point_exponential)
hist(fit_flash$residuals.sd,breaks = 100)
```


Still not work. So maybe for this dataset, after accounting for the Document size and background word frequency, there's no new multiplicative factor?

What if we subtract Document size and background word frequency from the latent data, and fit flash?

```{r}
y = fit$init_val$M_init - tcrossprod(cbind(l0),cbind(rep(1,p))) - tcrossprod(cbind(rep(1,n)),cbind(f0))
res = flash(y,ebnm.fn = ebnm::ebnm_point_exponential,var.type = 2,backfit = T,greedy.Kmax = 10)
```

```{r}
plot(svd(y)$d)
```


