---
title: "is vga binomial problem convex?"
author: "DongyueXie"
date: "2023-02-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

The evidence lower bound is,
\begin{equation}
    \begin{split}
        F(\bar\mu,v_\mu) =  x\bar\mu - n\E_q\log(1+\exp(\mu)) - \frac{\bar\mu^2+v_\mu - 2\bar\mu\beta}{2\sigma^2} + \frac{1}{2}\log v_\mu + \text{const}.
    \end{split}
\end{equation}

There is no close form for expected $\log(1+\exp(\mu))$ function. To evaluate $\E_q\log(1+\exp(\mu))$, we consider a Gauss-Hermite quadrature approximation as 
\begin{equation}
\begin{split}
       \E\log(1+\exp(\mu)) &= \int \frac{1}{\sqrt{\pi}}\exp(-\mu^2)\log(1+\exp(\sqrt{2v_\mu}\mu+\bar\mu))d\mu
       \\
       &\approx \frac{1}{\sqrt{\pi}} \sum_j w_j \log(1+\exp(\sqrt{2v_\mu}\mu_j+\bar\mu)),
\end{split}
\end{equation}
where $w_j,\mu_j$ are pre-selected fixed sampling points.


```{r}
source('code/binomial_mean/binomial_mean_splitting.R')
vga_binom_obj = function(m,lv,x,nb,beta,sigma2,gh_points){
  val = - sum(x*m - nb*Elog1pexp(m,exp(lv),gh_points) - (m^2+exp(lv)-2*m*beta)/2/sigma2 + lv/2)
  return(val)
}
```

```{r}
library(vebpm)
library(fastGHQuad)
library(ebnm)
library(Rfast)
```

```{r}
x = 1
nb = 5
beta=0
sigma2=1
n=1
n_gh = 10
gh_points = gaussHermiteData(n_gh)
lv = 0
m_list = seq(-3,3,length.out = 100)
val = c()
for(m in m_list){
  val  =c(val,vga_binom_obj(m,lv,x,nb,beta,sigma2,gh_points))
}
plot(m_list,val,type='l')
```

```{r}
x = 0
nb = 5
beta=0
sigma2=1
n=1
n_gh = 10
gh_points = gaussHermiteData(n_gh)
lv = 0
m_list = seq(-3,3,length.out = 100)
val = c()
for(m in m_list){
  val  =c(val,vga_binom_obj(m,lv,x,nb,beta,sigma2,gh_points))
}
plot(m_list,val,type='l')
```

```{r}
x = 0
nb = 1
beta=0
sigma2=1
n=1
n_gh = 10
gh_points = gaussHermiteData(n_gh)
lv = 0
m_list = seq(-3,3,length.out = 100)
val = c()
for(m in m_list){
  val  =c(val,vga_binom_obj(m,lv,x,nb,beta,sigma2,gh_points))
}
plot(m_list,val,type='l')
```


```{r}
x = 1
nb = 5
beta=0
sigma2=1
n=1
n_gh = 10
gh_points = gaussHermiteData(n_gh)
m = 0
v_list = seq(1e-5,3,length.out = 100)
val = c()
for(v in v_list){
  val  =c(val,vga_binom_obj(m,log(v),x,nb,beta,sigma2,gh_points))
}
plot(v_list,val,type='l')
```

```{r}
x = 1
nb = 5
beta=0
sigma2=0.5
n=1
n_gh = 10
gh_points = gaussHermiteData(n_gh)
m = 0
v_list = seq(1e-5,3,length.out = 100)
val = c()
for(v in v_list){
  val  =c(val,vga_binom_obj(m,log(v),x,nb,beta,sigma2,gh_points))
}
plot(v_list,val,type='l')
```
