---
title: "single cell library size distribution"
author: "DongyueXie"
date: "2022-12-08"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

I look at the distribution of single cell library size on two datasets.

The pbmc dataset.

```{r}
library(fastTopics)
library(Matrix)
data("pbmc_facs")
s = rowSums(pbmc_facs$counts)
hist(s,breaks = 100,main='pbmc data library size',probability = T)
```

Fit a gamma distribution on $s$, using method of moment.

```{r}
rate = mean(s)/var(s)
shape = mean(s)*rate
rate
shape
hist(rgamma(1e5,shape=shape,rate=rate),breaks = 100,main='simulated library size',probability = T)
```

Fit a inverse gamma distribution on $s$, using method of moment.
```{r}
library(invgamma)
shape = 2 + mean(s)^2/var(s)
rate = (shape-1)*mean(s)
shape
rate
hist(rinvgamma(1e5,shape=shape,rate=rate),breaks = 100,main='simulated library size',probability = T)
```

droplet data

```{r}
load('data/real_data_singlecell/droplet.RData')
```

```{r}
s = rowSums(counts)
hist(s,breaks = 100,main='droplet data library size',probability = T)
```

Fit a gamma distribution on $s$, using method of moment.

```{r}
rate = mean(s)/var(s)
shape = mean(s)*rate
rate
shape
hist(rgamma(1e5,shape=shape,rate=rate),breaks = 100,main='simulated library size',probability = T)
```

Fit a inverse gamma distribution on $s$, using method of moment.
```{r}
library(invgamma)
shape = 2 + mean(s)^2/var(s)
rate = (shape-1)*mean(s)
shape
rate
dd = rinvgamma(1e5,shape=shape,rate=rate)
hist(rinvgamma(1e5,shape=shape,rate=rate),breaks = 100,main='simulated library size',probability = T)
hist(dd[dd<max(s)],breaks = 100,main='simulated library size',probability = T)
```

## Conclusion

Gamma distribution seems not able to capture the peak of library size distribution. Inverse-gamma has heavier tails. I'll use inverse-Gamma, but set a maximum library size(truncate it).


