---
title: "log link benchmarking of Poisson means"
author: "DongyueXie"
date: "2022-11-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

We compare the methods with log-link on estimating the latent $\mu$ under the following simulation settings. We generate $n=1000$ samples from $y_j\sim \Poi(\exp(\mu_j))$, and $\mu_j$ are generated under the following different data-generating distributions. Each simulation was repeated for 50 times.

```{r}
library(vebpm)
library(ggplot2)
library(tidyverse)
source('code/poisson_mean/simulation_summary.R')
```

## $\mu_j \sim \pi_0\delta_0 + \pi_1N(0,2)$.

```{r}
out = readRDS('output/poisson_mean_simulation/poisson_mean/log_link50_n_1000_priormean_0_priorvar1_2.rds')
```

```{r}
get_summary_mean(out)
get_summary_mean_log(out,rm_method = 'ash_pois_log')
```

```{r}
plot(out$sim_data$log_Mean[1,],col='grey80',main='log_mean',ylab='')
plot(out$sim_data$Mean[1,],col='grey80',main='log_mean',ylab='')
```

## $\mu_j \sim \pi_0\delta_3 + \pi_1N(3,2)$

```{r}
out = readRDS('output/poisson_mean_simulation/poisson_mean/log_link50_n_1000_priormean_3_priorvar1_2.rds')
```

```{r}
get_summary_mean(out)
get_summary_mean_log(out)
```

## $\mu_j \sim \pi_0\delta_5 + \pi_1N(5,2)$

```{r}
out = readRDS('output/poisson_mean_simulation/poisson_mean/log_link50_n_1000_priormean_5_priorvar1_2.rds')
```

```{r}
get_summary_mean(out,rm_method = c('nb_lb','nb_pg'))
get_summary_mean_log(out,rm_method = c('nb_lb','nb_pg'))
```


## Why NB method performs bad?

Its performance is getting worse as x getting larger.

I set $r = 2*max(y)$. So sometimes $r$ is of order $10^5$, which is too large. We see from the results that setting $r$ large likely let veb algorithm get stuck at local optimum.

Let see 4 examples when NB methods perform worst of itself.

```{r}
par(mfrow=(c(2,2)))
for(i in c(1,21,23,24)){
  plot(out$sim_data$log_Mean[i,],col='grey80')
lines(out$output[[i]]$fitted_model$nb_lb$posterior$mean_log)
}
```

The corresponding $r$ are 

```{r}
for(i in c(1,21,23,24)){
  print(2*(max(out$sim_data$X[i,])))
}
```

Let's reduce $r$ to 1000 for the 1st simulation and see it improves

```{r}
for(i in c(1,21,23,24)){
  temp = nb_mean_lower_bound(out$sim_data$X[i,],r=1000)
 plot(out$sim_data$log_Mean[i,],col='grey80')
lines(temp$posterior$mean_log)
print(mse(out$sim_data$log_Mean[i,],temp$posterior$mean_log))
}
```

Further reduce $r$ to 100

```{r}
for(i in c(1,21,23,24)){
  temp = nb_mean_lower_bound(out$sim_data$X[i,],r=100)
 plot(out$sim_data$log_Mean[i,],col='grey80')
lines(temp$posterior$mean_log)
print(mse(out$sim_data$log_Mean[i,],temp$posterior$mean_log))
}
```


It seems that setting $r=1000$ is slightly better than $100$, but both are significantly better than $r\sim10^5$. 

## What happend to inversion penalization method?

Its performance is getting worse as x getting larger.

It seems that it has some convergence issues here. The estimated prior weights are still at the uniform initialization stage.

```{r}
out$output[[1]]$fitted_model$penalty_inversion$fitted_g
```

Check the gradient of posterior mean

```{r}
plot(vebpm:::f_obj_grad(out$output[[1]]$fitted_model$penalty_inversion$fit$optim_fit$par,out$sim_data$X[1,],out$output[[1]]$fitted_model$penalty_inversion$fitted_g$sd)[1:1000],ylab='gradient')
```

Check the gradient of prior weight

```{r}
plot(vebpm:::f_obj_grad(out$output[[1]]$fitted_model$penalty_inversion$fit$optim_fit$par,out$sim_data$X[1,],out$output[[1]]$fitted_model$penalty_inversion$fitted_g$sd)[1001:1032],ylab='gradient')
```

Check the gradient of prior mean

```{r}
plot(vebpm:::f_obj_grad(out$output[[1]]$fitted_model$penalty_inversion$fit$optim_fit$par,out$sim_data$X[1,],out$output[[1]]$fitted_model$penalty_inversion$fitted_g$sd)[1033],ylab='gradient')
```

It's not converging! Because the LBFGSB in `optim` uses relative tol(tol*f_value) for monitoring the convergence and the f_value is too large so iterations stop before convergence.

Need to increase number of iterations. 

```{r,eval=FALSE}
temp= pois_mean_penalized_inversion(out$sim_data$X[1,],tol=1e-10)
temp$fitted_g$weight
mse(temp$posterior$mean,out$sim_data$Mean[1,])
```

