---
title: "log link benchmark revised"
author: "DongyueXie"
date: "2022-11-09"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

See [previous](log_link_benchmarking.html) for an initial benchmarking results.

Fixed NB methods by reducing $r$ to 1000 or 100.

Fixed inversion convergence issue by 1. add const term to objective; 2. reduce the reltol

Fixed log1pexp so it can deal with large value.

Fixed VGA_optimize so that it does not give "LBFGSB needs finite fn value" warning, by skipping it.

Generated $n=1000$ observations from \[y_i\sim Poisson(\exp(\mu_i),\mu_i\sim g(\cdot).\]

Look at MSE of $\hat\mu$ and estimate of $\exp\mu$.

```{r}
library(vebpm)
library(ggplot2)
library(tidyverse)
source('code/poisson_mean/simulation_summary.R')
```

## $\mu_j \sim \pi_0\delta_0 + \pi_1N(0,2)$.

```{r}
out = readRDS('output/poisson_mean_simulation/poisson_mean/log_link31_n_1000_priormean_0_priorvar1_2.rds')
```

```{r}
get_summary_mean(out,rm_method = c('ash_pois_log', 'GMGM_pointmass'))
get_summary_mean_log(out,rm_method = c('ash_pois_log', 'GMGM_pointmass'))
```


## $\mu_j \sim \pi_0\delta_3 + \pi_1N(3,2)$

```{r}
out = readRDS('output/poisson_mean_simulation/poisson_mean/log_link31_n_1000_priormean_3_priorvar1_2.rds')
```

```{r}
get_summary_mean(out,rm_method = c('ash_pois_log', 'GMGM_pointmass'))
get_summary_mean_log(out,rm_method = c('ash_pois_log', 'GMGM_pointmass'))
```

## $\mu_j \sim \pi_0\delta_5 + \pi_1N(5,2)$

```{r}
out = readRDS('output/poisson_mean_simulation/poisson_mean/log_link31_n_1000_priormean_5_priorvar1_2.rds')
```

```{r}
get_summary_mean(out,rm_method = c('ash_pois_log', 'GMGM_pointmass'))
get_summary_mean_log(out,rm_method = c('ash_pois_log', 'GMGM_pointmass'))
```
