---
title: "log(Mixture of normal density) concavity and variance ratio"
author: "Dongyue Xie"
date: "2022-06-06"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Look at log of density function 

\[g(x) = \sum_k w_k N(x;0,\sigma^2_k).\]



```{r}
#'second derivative of log g(x) where g() is a mixture of Gaussian
#'@param x observation, a scalar.
#'@param w mixture weights.
#'@param s2 variances of mixture components.
lg_d2 = function(x,w,s2){
  n1 = sum(w/sqrt(2*pi*s2)*exp(-x^2/2/s2)*((x/s2)^2-1/s2))
  d = sum(w/sqrt(2*pi*s2)*exp(-x^2/2/s2))
  n2 = -sum(w/sqrt(2*pi*s2)*exp(-x^2/2/s2)*x/s2)
  return(n1/d - (n2/d)^2)
}

#'second derivative of log g(x) where g() is a mixture of Gaussian
#'x is a vector
lg_d2_vec = function(x,w,s2){
  n = length(x)
  out = rep(0,n)
  for(i in 1:n){
    out[i] = lg_d2(x[i],w,s2)
  }
  return(out)
}

#'density of log g(x)
#'x is a scalar
lg = function(x,w,s2){
  return(log(sum(w/sqrt(2*pi*s2)*exp(-x^2/2/s2))))
}

#'density of log g(x)
#'x is a vector
lg_vec = function(x,w,s2){
  n = length(x)
  out = rep(0,n)
  for(i in 1:n){
    out[i] = lg(x[i],w,s2)
  }
  return(out)
}

plot_lg = function(x,w,s2){
  plot(x,lg_vec(x,w,s2),type='l',xlab='x',ylab='log(g)',
       main=paste('w=(',paste(w,collapse = ", "),'); s2=(',paste(s2,collapse = ", "),')',sep=''))
}

plot_lg_d2 = function(x,w,s2){
  plot(x,lg_d2_vec(x,w,s2),type='l',xlab='x',ylab="(log g)''",
       main=paste('w=(',paste(w,collapse = ", "),'); s2=(',paste(s2,collapse = ", "),')',sep=''))
}
```

## log concave and variance ratio

Fix $w=(0.5,0.5)$, and smaller variance to be 1. Then vary the larger one from 1.1 to 4. Check $max(\log''(g))$. If it's smaller than 0, the $\log g$ is concave. Then we divide the both variances by 10, so the smaller one is 0.1, and plot the same figure. Finally, we multiply both variances by 10, so the smaller variance is 10, and plot the same figure.

```{r,fig.width=9,fig.height=3}
x = seq(-10,10,length.out = 1000)
w = c(0.5,0.5)
s2_list=seq(1.1,4,by=0.1)
out = c()
for(ss in s2_list){
  out = c(out,max(lg_d2_vec(x,w,c(1,ss))))
}

par(mfrow=c(1,3))

plot(s2_list,out,type='l',xlab='ratio',ylab="max(log''(g))")
abline(h=0,lty=2)


s2_list=seq(1.1,4,by=0.1)/10
out = c()
for(ss in s2_list){
  out = c(out,max(lg_d2_vec(x,w,c(0.1,ss))))
}
plot(s2_list,out,type='l',xlab='ratio',ylab="max(log''(g))")
abline(h=0,lty=2)

s2_list=seq(1.1,4,by=0.1)*10
out = c()
for(ss in s2_list){
  out = c(out,max(lg_d2_vec(x,w,c(10,ss))))
}
plot(s2_list,out,type='l',xlab='ratio',ylab="max(log''(g))")
abline(h=0,lty=2)
```

All lines have the same shape and the intersection points are at the same position. This confirms that with weights fixed, the concavaity is related to the variance ratios.


## concavity of the penalty

Now consider the problem $y|\theta\sim N(\theta,1)$, $\theta\sim 0.5\delta_0+0.5N(0,2)$. What does the penalties $\rho(\theta)$ and $\rho(S(\theta))$ look like?

\[c(\theta) = \rho(S(\theta)) = -l_{NM}(\theta;g,s^2)-\frac{(\theta-S_{g,s}(\theta))^2}{2s^2}\]

To evaluate $\rho(\theta)$, we find $z$ such that $S(z) = \theta$ then $\rho(\theta) = \rho(S(z))$.

To determine the convexity, we evaluate the second derivative of $\rho(\cdot)$:

For the compound penalty $c(\theta) = -l(\theta) - \frac{s^2(l'(\theta))^2}{2}$

\[c'(\theta) = -l'(\theta) - s^2l'(\theta)l''(\theta)\]

\[c''(\theta) = -l''(\theta) - s^2(l''(\theta)^2+l'''(\theta)l'(\theta))\]

For the original penalty:

\[\rho''(S(\theta)) = -l''(\theta)/S'(\theta).\]

\[S(\theta) = \theta + s^2l_{NM}'(\theta)\]

\[S'(\theta) = 1 + s^2l_{NM}''(\theta)\]

\[\rho''(S(\theta)) = -l''(\theta)/S'(\theta) = -l''(\theta)/(1+s^2l''(\theta)).\]

To evaluate $\rho''(\theta)$, we find $z$ such that $S(z) = \theta$ then $\rho''(\theta) = \rho''(S(z))$.

```{r}
#'density of log g(x)
#'x is a scalar
lg = function(x,w,s2){
  return(log(sum(w*dnorm(x,0,sd=sqrt(s2)))))
}

#'density of log g(x)
#'x is a vector
lg_vec = function(x,w,s2){
  n = length(x)
  out = rep(0,n)
  for(i in 1:n){
    out[i] = lg(x[i],w,s2)
  }
  return(out)
}
```

```{r}
library(ashr)
library(ebnm)
#g = normalmix(pi = c(0,w),mean=rep(0,K+1),sd = c(0,sqrt(s2)))


nm = function(z,g,s){
  fit = ash(z,s,g = g,fixg = T)
  return(list(l_nm = fit$loglik, pm=fit$result$PosteriorMean))
}

penalty = function(theta,g,s){
  temp = nm(theta,g,s)
  return(-temp$l_nm - (theta-temp$pm)^2/2/s^2)
}

penalty_vec = function(x,g,s){
  n = length(x)
  out = rep(0,n)
  for(i in 1:n){
    out[i] = penalty(x[i],g,s)
  }
  return(out)
}




#'posterior mean operator
S = function(z,g,s){
  pw = log(g$pi)+dnorm(z,mean=0,sd=sqrt(s^2+g$sd^2),log=TRUE)
  pw = pw - max(pw)
  pw = exp(pw)/sum(exp(pw))
  pm = z/(1+s^2/g$sd^2)
  return(sum(pw*pm))
}

#'Solve S(z) = theta for a given theta
#'@param theta 
#'@param g a normalmix object with weights pi and sd
#'@param s known standard error
inv_S = function(theta,g,s=1){
  eqn = function(z,theta,g,s){
    S(z,g,s) - theta
  }
  if(theta==0){
    sol = 0
  }else if(theta>0){
    sol = uniroot(eqn,interval=c(0,theta/min(1/(1+s^2/(g$sd^2+.Machine$double.eps)))),theta=theta,g=g,s=s)$root
  }else{
    sol = uniroot(eqn,interval=c(theta/min(1/(1+s^2/(g$sd^2+.Machine$double.eps))),0),theta=theta,g=g,s=s)$root
  }
  return(sol)
}

inv_S_vec = function(x,g,s=1){
  n = length(x)
  out = c()
  for(i in 1:n){
    out[i] = inv_S(x[i],g,s)
  }
  out
}

```

```{r}
plot_penalty = function(x,g,s=1){
  
  
  # plot compound penalty
  compound_penalty = penalty_vec(x,g,s)
  z = inv_S_vec(x,g,s)
  original_penalty = penalty_vec(z,g,s)
  
  plot(x,compound_penalty,type='l',xlab='theta',ylab='penalty',ylim=range(c(compound_penalty,original_penalty)),
       main=paste('w=(',paste(g$pi,collapse = ", "),'); sd=(',paste(round(g$sd,2),collapse = ", "),')',sep=''))
  lines(x,original_penalty,col='blue',lty=2)
  legend('bottomright',c(expression(rho(S(theta))),expression(rho(theta))),lty=c(1,2),col=c(1,'blue'))
}

f = expression(log(w1/sqrt((s1_2+s2)*2*pi)*exp(-x^2/2/(s1_2+s2))+w2/sqrt((s2_2+s2)*2*pi)*exp(-x^2/2/(s2_2+s2))))
f1 = D(f,'x')
f2 = D(f1,'x')
f3 = D(f2,'x')

lf_d3 = function(x = seq(-10,10,length.out = 100),
                     s = 1,
                     w1 = 1,
                     s1_2 = 0.1,
                     s2_2 = 2){
  s2 = s^2
  w2 = 1-w1
  y = eval(f3)
  y
}

lf_d2 = function(x = seq(-10,10,length.out = 100),
                     s = 1,
                     w1 = 1,
                     s1_2 = 0.1,
                     s2_2 = 2){
  s2 = s^2
  w2 = 1-w1
  y = eval(f2)
  y
}

lf_d1 = function(x = seq(-10,10,length.out = 100),
                     s = 1,
                     w1 = 1,
                     s1_2 = 0.1,
                     s2_2 = 2){
  s2 = s^2
  w2 = 1-w1
  y = eval(f1)
  y
}

compound_penalty_d2 = function(x,g,s){
  -lf_d2(x,s,g$pi[1],g$sd[1]^2,g$sd[2]^2) - s^2*(lf_d2(x,s,g$pi[1],g$sd[1]^2,g$sd[2]^2)^2+lf_d3(x,s,g$pi[1],g$sd[1]^2,g$sd[2]^2)*lf_d1(x,s,g$pi[1],g$sd[1]^2,g$sd[2]^2))
}

original_penalty_d2 = function(x,g,s){
  -lf_d2(x,s,g$pi[1],g$sd[1]^2,g$sd[2]^2)/(1+s^2*lf_d2(x,s,g$pi[1],g$sd[1]^2,g$sd[2]^2))
}

plot_penalty_d2 = function(x,g,s=1){
  compound_penalty = compound_penalty_d2(x,g,s)
  z = inv_S_vec(x,g,s)
  original_penalty = original_penalty_d2(z,g,s)
  plot(x,compound_penalty,type='l',xlab='theta',ylab='penalty',ylim=range(c(compound_penalty,original_penalty)),
       main=paste('w=(',paste(g$pi,collapse = ", "),'); sd=(',paste(round(g$sd,2),collapse = ", "),')',sep=''))
  lines(x,original_penalty,col='blue',lty=2)
  legend('topright',c(expression(rho(S(theta))),expression(rho(theta))),lty=c(1,2),col=c(1,'blue'))
}

```

```{r}
library(ashr)
x = seq(-10,10,length.out = 101)
g = normalmix(pi = c(0.5,0.5),mean=rep(0,2),sd = c(0,sqrt(2)))
plot_penalty(x,g)
plot_penalty_d2(x,g)
```

