---
title: "multiplicative vs additive PMF"
author: "DongyueXie"
date: "2023-08-17"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

I want to find a simple example to illustrate the difference between multiplicative and additive methods.

The multiplicative effects come with the log link in Poisson model can be understood as follows:

Let $\boldsymbol{x}_i$ denote the a data vector, then we can write $\mathbb{E}(\boldsymbol{x}_i) = \exp(\boldsymbol{\mu} + \sum_kl_{ik}\boldsymbol{f}_k)=\exp(\boldsymbol{\mu})\prod_k \exp(l_{ik}\boldsymbol{f}_k).$

Here I start with a very simple example where the two factors look like:

$\boldsymbol{f}_1 = (1,1,1,1,0,0,0,0), \boldsymbol{f}_2=(1,1,0,0,1,1,0,0)$.

So for the $i$th sample, its mean is $background\times \exp(l_{i1}\boldsymbol{f}_1)\times \exp(l_{i2}\boldsymbol{f}_2)$.

WLOG let's set background to be all 1's. The observations(elements) of $i$th sample would be $\exp(l_{i1})\exp(l_{i2}),\exp(l_{i1}),\exp(l_{i2}),1$.

Let's set the following loadings:

1. $l_1=l_2=0$; 2. $l_1=2, l_2=0$; 3. $l_1=0,l_2=2$; 4. $l_1=4,l_2=2$.

The corresponding Poisson rates are 

$(1,1,1,1),(\exp(2),\exp(2),1,1),(\exp(2),1,\exp(2),1), (\exp(6),\exp(4),\exp(2),1)$.



```{r}
N = 100
Ftrue = cbind(c(1,1,1,1,0,0,0,0),c(1,1,0,0,1,1,0,0))
set.seed(12345)
#Ltrue = matrix(rexp(N*2,1),nrow=N)
Ltrue = cbind(rep(c(0,2,0,4),each=N/4),rep(c(0,0,2,2),each=N/4))
mu = tcrossprod(Ltrue,Ftrue)
lambda = exp(mu)
Y = matrix(rpois(N*nrow(Ftrue),lambda),nrow=N)

Ftrue
plot(Ltrue[,1])
plot(Ltrue[,2])
```

Can NMF recover Ltrue and Ftrue, if we directly apply NMF on $LF$?

```{r}
nmf_fit = NNLM::nnmf(mu,k=2,loss='mse')
plot(nmf_fit$H[1,])
plot(nmf_fit$H[2,])

```

Looks good.

How about running EBNMF on $LF$? It is not working because "The data matrix must not have any rows or columns whose entries are either identically zero or all missing."

```{r,eval=F}
ebnmf_fit = flashier::flash(mu,ebnm_fn = c(ebnm::ebnm_unimodal_nonnegative,ebnm::ebnm_unimodal_nonnegative),var_type = 0,backfit = T)
```

Now we fit NMF to the data matrix Y:

Let's start with k=2

```{r}
fit0 = fastTopics::fit_poisson_nmf(Y,2)
plot(fit0$F[,1])
plot(fit0$F[,2])

plot(fit0$L[,1])
plot(fit0$L[,2])
```

Then k = 3

```{r}
fit0 = fastTopics::fit_poisson_nmf(Y,3)
plot(fit0$F[,1])
plot(fit0$F[,2])
plot(fit0$F[,3])

plot(fit0$L[,1])
plot(fit0$L[,2])
plot(fit0$L[,3])
```

Now we try to fit ebnmf on log transformed data:

```{r}
fit2 = flashier::flash(log(1+Y),ebnm_fn = c(ebnm::ebnm_point_exponential,ebnm::ebnm_point_exponential),var_type = 0,backfit = T)
plot(fit2$F_pm[,1])
plot(fit2$F_pm[,2])
plot(fit2$F_pm[,3])

plot(fit2$L_pm[,1])
plot(fit2$L_pm[,2])
plot(fit2$L_pm[,3])
```

```{r}
fit2 = flashier::flash(log(0.5+Y),ebnm_fn = c(ebnm::ebnm_point_exponential,ebnm::ebnm_point_exponential),var_type = 0,backfit = T)
plot(fit2$F_pm[,1])
plot(fit2$F_pm[,2])
```

```{r}
fit1 = ebpmf::ebpmf_log(Y,l0=0,f0=0,var_type = 'constant',
                        flash_control = list(fix_l0=T,fix_f0=T,ebnm.fn=c(ebnm::ebnm_unimodal_nonnegative,ebnm::ebnm_unimodal_nonnegative),
                                                                         loadings_sign=1,factors_sign =1))
```


