---
title: "Normal Mean problem Penalty"
author: "Dongyue Xie"
date: "2022-05-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

What do the penalties $\rho_{g,s^2}(\theta)$ and $c_{g,s^2}(\theta) = \rho_{g,s^2}(S_{g,s^2}(\theta))$  look like? 

The first one is the original penalty and we'll need inversion $S^{-1}(\cdot)$ to evaluate it; the second one is the compound penalty.

\[\rho_{g,s^2}(\theta) = \rho_{g,s^2}(S_{g,s^2}(S^{-1}_{g,s^2}(\theta))) = c_{g,s^2}(S^{-1}_{g,s^2}(\theta))\]

\[c_{g,s^2}(\theta) = \rho_{g,s^2}(S_{g,s^2}(\theta)) = -l_{NM}(\theta;g,s^2)-\frac{(\theta-S_{g,s}(\theta))^2}{2s^2}\]

(To evaluate $\rho(\theta)$, we find $z$ such that $S(z) = \theta$ then $\rho(\theta) = c(z)$.)

We also evaluate the second derivative of the penalties:

For the compound penalty $c(\theta) = -l(\theta) - \frac{s^2(l'(\theta))^2}{2}$

\[c'(\theta) = -l'(\theta) - s^2l'(\theta)l''(\theta)\]

\[c''(\theta) = -l''(\theta) - s^2(l''(\theta)^2+l'''(\theta)l'(\theta))\]

For the penalty $\rho(\theta)$:

\[\rho_{g,s^2}'(\theta) = \frac{S^{-1}_{g,s^2}(\theta)-\theta}{s^2}\]

\[\rho_{g,s^2}''(\theta) = \frac{1}{s^2}\left(\frac{1}{1+s^2l''_{nm}(S^{-1}_{g,s^2}(\theta))}-1\right) = \frac{-l''_{nm}(S^{-1}_{g,s^2}(\theta))}{1+s^2l''_{nm}(S^{-1}_{g,s^2}(\theta))}\]

\[\rho''(S(\theta)) = -l''(\theta)/S'(\theta).\]

\[S(\theta) = \theta + s^2l_{NM}'(\theta)\]

\[S'(\theta) = 1 + s^2l_{NM}''(\theta)\]

\[\rho''(S(\theta)) = -l''(\theta)/S'(\theta) = -l''(\theta)/(1+s^2l''(\theta)).\]



```{r}
source('code/normal_mean_model_utils.R')

# nm = function(z,g,s){
#   fit = ash(z,s,g = g,fixg = T)
#   return(list(l_nm = fit$loglik, pm=fit$result$PosteriorMean))
# }
# 
# penalty = function(theta,g,s){
#   temp = nm(theta,g,s)
#   return(-temp$l_nm - (theta-temp$pm)^2/2/s^2)
# }
# 
# penalty_vec = function(x,g,s){
#   n = length(x)
#   out = rep(0,n)
#   for(i in 1:n){
#     out[i] = penalty(x[i],g,s)
#   }
#   return(out)
# }
# 
# 
# 
# 
# #'posterior mean operator
# S = function(z,g,s){
#   pw = log(g$pi)+dnorm(z,mean=0,sd=sqrt(s^2+g$sd^2),log=TRUE)
#   pw = pw - max(pw)
#   pw = exp(pw)/sum(exp(pw))
#   pm = z/(1+s^2/g$sd^2)
#   return(sum(pw*pm))
# }
 
# #'Solve S(z) = theta for a given theta
# #'@param theta 
# #'@param g a normalmix object with weights pi and sd
# #'@param s known standard error
# inv_S = function(theta,g,s=1){
#   eqn = function(z,theta,g,s){
#     S(z,g,s) - theta
#   }
#   if(theta==0){
#     sol = 0
#   }else if(theta>0){
#     sol = uniroot(eqn,interval=c(0,theta/min(1/(1+s^2/g$sd^2))),theta=theta,g=g,s=s)$root
#   }else{
#     sol = uniroot(eqn,interval=c(theta/min(1/(1+s^2/g$sd^2)),0),theta=theta,g=g,s=s)$root
#   }
#   return(sol)
# }
# 
# inv_S_vec = function(x,g,s=1){
#   n = length(x)
#   out = c()
#   for(i in 1:n){
#     out[i] = inv_S(x[i],g,s)
#   }
#   out
# }

library(vebpm)
plot_penalty = function(x,s=1,w,mu=0,grid){

  # plot compound penalty
  compound_penalty = nm_penalty_compound(x,s,w,mu,grid)
  original_penalty = nm_penalty(x,s,w,mu,grid)
  
  plot(x,compound_penalty,type='l',xlab='theta',ylab='penalty',ylim=range(c(compound_penalty,original_penalty)),
       main=(paste(pi,'=(',paste(w,collapse = ", "),'); ',sigma,'=(',paste(grid,collapse = ", "),')',sep='')))
  lines(x,original_penalty,col='blue',lty=2)
  legend('bottomright',c(expression(rho(S(theta))),expression(rho(theta))),lty=c(1,2),col=c(1,'blue'))
}

plot_penalty_hess = function(x,s=1,w,grid){

  # plot compound penalty
  compound_penalty = nm_penalty_compound_hess(x,s,w,grid)
  original_penalty = nm_penalty_hess(x,s,w,grid)
  
  plot(x,compound_penalty,type='l',xlab='theta',ylab='2nd derivative of penalty',ylim=range(c(compound_penalty,original_penalty)),
       main=paste('w=(',paste(w,collapse = ", "),'); grid=(',paste(grid,collapse = ", "),')',sep=''))
  lines(x,original_penalty,col='blue',lty=2)
  legend('topright',c(expression(rho^"''"*(S(theta))),expression(rho^"''"*(theta))),lty=c(1,2),col=c(1,'blue'))
}

library(ggplot2)
```

```{r,fig.height=4,fig.width=4}
n = 101
x = seq(-10,10,length.out = n)
s = rep(1,n)
w = c(0.9,0.1)
par(mfrow=c(2,2))
grid_list=c(1,2,3,5)
# for(gg in grid_list){
#   plot_penalty(x,s,w,0,c(0,gg))
# }


gg = 1
compound_penalty = nm_penalty_compound(x,s,w,mu=0,c(0,gg))
original_penalty = nm_penalty(x,s,w,mu=0,c(0,gg))



# Create data frame with x, compound_penalty, and original_penalty columns
df <- data.frame(x, compound_penalty, original_penalty)

# Create the ggplot with type='l' (lines)
p1 = ggplot(df, aes(x=x)) + 
  geom_line(aes(y=compound_penalty), color="black", linetype="solid") +
  geom_line(aes(y=original_penalty), color="blue", linetype="dashed") +
  
  # Add x and y labels
  xlab(expression(theta)) + ylab("penalty") + 
  
  # Set the y-axis limits to range of compound_penalty and original_penalty
  ylim(range(c(df$compound_penalty, df$original_penalty))) +
  
  # Add title
  ggtitle(expression(paste(pi,'=(0.9, 0.1); ',sigma,'=(0, 1)',sep='')))

```

```{r,fig.height=4,fig.width=4}
n = 101
x = seq(-10,10,length.out = n)
s = rep(1,n)
w = c(0.9,0.1)
par(mfrow=c(2,2))
grid_list=c(1,2,3,5)
# for(gg in grid_list){
#   plot_penalty(x,s,w,0,c(0,gg))
# }


gg = 2
compound_penalty = nm_penalty_compound(x,s,w,mu=0,c(0,gg))
original_penalty = nm_penalty(x,s,w,mu=0,c(0,gg))



# Create data frame with x, compound_penalty, and original_penalty columns
df <- data.frame(x, compound_penalty, original_penalty)

# Create the ggplot with type='l' (lines)
p2 = ggplot(df, aes(x=x)) + 
  geom_line(aes(y=compound_penalty), color="black", linetype="solid") +
  geom_line(aes(y=original_penalty), color="blue", linetype="dashed") +
  
  # Add x and y labels
  xlab(expression(theta)) + ylab("penalty") + 
  
  # Set the y-axis limits to range of compound_penalty and original_penalty
  ylim(range(c(df$compound_penalty, df$original_penalty))) +
  
  # Add title
  ggtitle(expression(paste(pi,'=(0.9, 0.1); ',sigma,'=(0, 2)',sep='')))

```


```{r,fig.height=4,fig.width=4}
n = 101
x = seq(-10,10,length.out = n)
s = rep(1,n)
w = c(0.5,0.5)
par(mfrow=c(2,2))
grid_list=c(1,2,3,5)
# for(gg in grid_list){
#   plot_penalty(x,s,w,0,c(0,gg))
# }


gg = 1
compound_penalty = nm_penalty_compound(x,s,w,mu=0,c(0,gg))
original_penalty = nm_penalty(x,s,w,mu=0,c(0,gg))



# Create data frame with x, compound_penalty, and original_penalty columns
df <- data.frame(x, compound_penalty, original_penalty)

# Create the ggplot with type='l' (lines)
p3 = ggplot(df, aes(x=x)) + 
  geom_line(aes(y=compound_penalty), color="black", linetype="solid") +
  geom_line(aes(y=original_penalty), color="blue", linetype="dashed") +
  
  # Add x and y labels
  xlab(expression(theta)) + ylab("penalty") + 
  
  # Set the y-axis limits to range of compound_penalty and original_penalty
  ylim(range(c(df$compound_penalty, df$original_penalty))) +
  
  # Add title
  ggtitle(expression(paste(pi,'=(0.5, 0.5); ',sigma,'=(0, 1)',sep='')))

```

```{r,fig.height=4,fig.width=4}
n = 101
x = seq(-10,10,length.out = n)
s = rep(1,n)
w = c(0.5,0.5)
par(mfrow=c(2,2))
grid_list=c(1,2,3,5)
# for(gg in grid_list){
#   plot_penalty(x,s,w,0,c(0,gg))
# }


gg = 2
compound_penalty = nm_penalty_compound(x,s,w,mu=0,c(0,gg))
original_penalty = nm_penalty(x,s,w,mu=0,c(0,gg))



# Create data frame with x, compound_penalty, and original_penalty columns
df <- data.frame(x, compound_penalty, original_penalty)

# Create the ggplot with type='l' (lines)
p4 = ggplot(df, aes(x=x)) + 
  geom_line(aes(y=compound_penalty), color="black", linetype="solid") +
  geom_line(aes(y=original_penalty), color="blue", linetype="dashed") +
  
  # Add x and y labels
  xlab(expression(theta)) + ylab("penalty") + 
  
  # Set the y-axis limits to range of compound_penalty and original_penalty
  ylim(range(c(df$compound_penalty, df$original_penalty))) +
  
  # Add title
  ggtitle(expression(paste(pi,'=(0.5, 0.5); ',sigma,'=(0, 2)',sep='')))

```

```{r}
library(gridExtra)
grid.arrange(p1,p2,p3,p4,nrow=2)
```

```{r,fig.height=12,fig.width=12}
n = 101
x = seq(-10,10,length.out = n)
s = rep(1,n)
w = c(0.5,0.5)
par(mfrow=c(2,2))
grid_list=c(1,2,3,5)
for(gg in grid_list){
  plot_penalty(x,s,w,c(0,gg))
}

for(gg in grid_list){
  plot_penalty_hess(x,s,w,c(0,gg))
}

```


```{r,fig.height=12,fig.width=12}
n = 101
x = seq(-10,10,length.out = n)
s = rep(1,n)
w = c(0.1,0.9)
par(mfrow=c(2,2))
grid_list=c(1,2,3,5)
for(gg in grid_list){
  plot_penalty(x,s,w,c(0,gg))
}

for(gg in grid_list){
  plot_penalty_hess(x,s,w,c(0,gg))
}

```


```{r,eval=F}
#'softmax
softmax = function(a){
  exp(a-max(a))/sum(exp(a-max(a)))
}
s2 = exp(seq(-8,5,by=log(2)))
w = softmax(log(1/s2))
round(w,2)
round(s2,2)
K = length(w)
g = normalmix(pi = w,mean=rep(0,K),sd = sqrt(s2))
x = seq(-10,10,length.out = 101)
plot_penalty(x,w,s,sqrt(s2))
```

