---
title: "Normal mean penalized form optimization"
author: "Dongyue Xie"
date: "2022-05-25"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

\[y|\mu\sim N(\mu,s^2),\mu\sim g(\cdot)\]

Optimization: $\theta = E_q\mu$,

\[\min_{\theta,g}h(\theta,g) = \frac{1}{2s^2}(y-\theta)^2+\rho_{g,s}(\theta).\]


\[S_{g,s^2}(z) = z +s^2l'_{NM}(z;g,s^2)\]


The penalty term is only tractable at $\rho(S(\cdot))$. There are two ways to accommodate this. 

Define the penalty evaluated at $S_{g,s^2}(\cdot)$ as $c_{g,s^2}(\cdot):=\rho_{g,s^2}(S_{g,s^2}(\cdot))$.

1. Compound way


Let $\theta = S_{g,s^2}(z)$ such that $\rho_{g,s^2}(\theta) = c_{g,s^2}(z)$, the optimization problem is 


\[\min_{z,g} \tilde h(z,g) = \frac{1}{2s^2}(y-S_{g,s^2}(z))^2-l_{NM}(z;g,s^2)-(z-S_{g,s^2}(z))^2/(2s^2)\]

\[\min_{z,g} \tilde h(z,g) = \frac{1}{2s^2}(y-z-s^2l'_{NM}(z;g,s^2))^2-l_{NM}-s^2(l'_{NM})^2/2\]


Then set $\hat\theta = S(z)$. 

2. Inversion way


Note that $c_{g,s^2}(S^{-1}_{g,s^2}(\theta)) = \rho_{g,s^2}(\theta)$, so we can write the optimization problem as 

\[\min_{\theta,g}  h(\theta,g) = \frac{1}{2s^2}(y-\theta)^2-c_{g,s^2}(S^{-1}_{g,s^2}(\theta))\]

For the inversion method implementation and illustration, see [here](normal_mean_penalized_optimization_inversion.html)



```{r}
library(vebpm)
```

generate data, and get grid by running ash

```{r,warning=FALSE}
set.seed(12345)
n = 200
w0 = 0.9
mu = c(rep(0,round(n*w0)),rep(10,n-round(n*w0)))
w_true = c(w0,1-w0)
grid_true = c(0.01,7)
s = rep(1,n)
y = rnorm(n,mu,s)
library(ashr)
grids = ebnm:::get_ashr_grid(y,s)

system.time(fit.ash <- ash(y,s,mixcompdist = 'normal',pointmass=FALSE,prior='uniform',mixsd=grids))
system.time(fit_inv <- ebnm_penalized_inversion(y,s,g_init = list(sd=grids)))
system.time(fit_compound <- ebnm_penalized_compound(y,s,g_init = list(sd=grids)))
```

```{r}

library(ggplot2)
library(gridExtra)

# plot 1
p1 <- ggplot() + 
  geom_point(aes(x = 1:length(y), y = y), col = 'grey80', pch = 20) + 
  geom_line(aes(x = 1:length(mu), y = mu), col = 'grey60') + 
  geom_line(aes(x = 1:length(fit.ash$result$PosteriorMean), y = fit.ash$result$PosteriorMean)) + 
  labs(title = 'ash', x = '', y = '') +
  scale_color_manual(values = c('grey80', 'grey60', 'black')) +
  guides(colour = guide_legend(override.aes = list(pch = c(20, NA, NA), lty = c(NA, 1, 1)))) +
  theme_bw()

# plot 2
p2 <- ggplot() + 
  geom_point(aes(x = 1:length(y), y = y), col = 'grey80', pch = 20) + 
  geom_line(aes(x = 1:length(mu), y = mu), col = 'grey60') + 
  geom_line(aes(x = 1:length(fit_inv$posterior$mean), y = fit_inv$posterior$mean)) + 
  labs(title = 'inversion', x = '', y = '') +
  scale_color_manual(values = c('grey80', 'grey60', 'black')) +
  guides(colour = guide_legend(override.aes = list(pch = c(20, NA, NA), lty = c(NA, 1, 1)))) +
  theme_bw()

# plot 3
p3 <- ggplot() + 
  geom_point(aes(x = 1:length(y), y = y), col = 'grey80', pch = 20) + 
  geom_line(aes(x = 1:length(mu), y = mu), col = 'grey60') + 
  geom_line(aes(x = 1:length(fit_compound$posterior$mean), y = fit_compound$posterior$mean)) + 
  labs(title = 'compound', x = '', y = '') +
  scale_color_manual(values = c('grey80', 'grey60', 'black')) +
  guides(colour = guide_legend(override.aes = list(pch = c(20, NA, NA), lty = c(NA, 1, 1)))) +
  theme_bw()

# arrange the plots
grid.arrange(p1, p2, p3, ncol = 1)

```

```{r}
fit.ash$fitted_g$pi
fit_inv$fitted_g$pi
fit_compound$fitted_g$pi
```

