---
title: "Normal mean penalized form optimization - Inversion method"
author: "Dongyue Xie"
date: "2022-09-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

\[y|\mu\sim N(\mu,s^2),\mu\sim g(\cdot)\]

Optimization: $\theta = E_q\mu$,

\[\min_{\theta,g}h(\theta,g) = \frac{1}{2s^2}(y-\theta)^2+\rho_{g,s}(\theta).\]

We use inversion method for solving the optimization problem. 

For another method, see [compound method](normal_mean_penalized_optimization.html)

```{r}
source("code/normal_mean_model_utils.R")
```

generate data, and get grid by running ash

```{r,warning=FALSE}
set.seed(12345)
n = 200
w0 = 0.9
mu = c(rep(0,round(n*w0)),rep(10,n-round(n*w0)))
w_true = c(w0,1-w0)
grid_true = c(0.01,7)
s = rep(1,n)
y = rnorm(n,mu,s)
library(ashr)
fit.ash = ashr::ash(y,s,mixcompdist = 'normal')
#grid = exp(seq(log(s/100),log(sqrt(max(abs(y^2-s^2)))),by=log(sqrt(2))))
#fit.ash = S(y,s,w_true,grid_true)
#plot(fit.ash$fitted_g$sd,fit.ash$fitted_g$pi)
grid = fit.ash$fitted_g$sd
grid = grid[-1]
K = length(grid)

#plot(y,main='ash fit',col='grey80')
#lines(mu,col='grey60')
#lines(fit.ash$result$PosteriorMean)
#legend('topleft',c('data','true mean','ash posteriorMean'),pch=c(1,NA,NA),lty=c(NA,1,1),col=c('grey80','grey60',1))
```

## Known prior

```{r}
f_obj_known_g = function(theta,y,s,w,grid,z_range){
  z = S_inv(theta,s,w,grid,z_range)
  return((y-theta)^2/2/s^2 - l_nm(z,s,w,grid)-(z-theta)^2/2/s^2)
}

f_obj_known_g_grad = function(theta,y,s,w,grid,z_range){
  z = S_inv(theta,s,w,grid,z_range)
  return((z-y)/s^2)
}

ebnm_penalized_inversion_known_g = function(x,s,w,grid,theta_init = NULL,z_range=NULL,opt_method = 'L-BFGS-B'){
  n = length(x)
  
  if(length(s)==1){
    s = rep(s,n)
  }
  if(is.null(theta_init)){
    theta_init = rep(0,n)
  }
  if(is.null(z_range)){
    z_range = range(x) + c(-1,1)
  }
  theta = double(n)
  for(i in 1:n){
    theta[i] = optim(theta_init[i],
                     fn=f_obj_known_g,
                     gr = f_obj_known_g_grad,
                   y=x[i],
                   s=s[i],
                   w=w,
                   grid=grid,
                   z_range=z_range,
                   method = opt_method)$par
  }
  return(theta)
}
```      

```{r}
ploter = function(fit,y,mu,main='known prior'){
  plot(y,main=main,col='grey80')
  lines(mu,col='grey60')
  lines(fit)
  legend('topleft',c('data','true mean','estimated mean'),pch=c(1,NA,NA),lty=c(NA,1,1),col=c('grey80','grey60',1))
}
```

### Init at data $y$.

```{r}
fit = ebnm_penalized_inversion_known_g(y,s,w_true,grid_true,theta_init = y,opt_method = 'L-BFGS-B')
```

```{r}
ploter(fit,y,mu,main='known prior, init at y')
```


### Init at 0.

```{r}
fit = ebnm_penalized_inversion_known_g(y,s,w_true,grid_true,theta_init = rep(0,n),opt_method = 'L-BFGS-B')
```

```{r}
ploter(fit,y,mu,main='known prior, init at 0')
```


## Estimate prior

```{r}
#'objective function
#'@param theta (theta,w)
#'@param grid prior sds
f_obj = function(params,y,s,grid,z_range,opt_method='L-BFGS-B'){
  n = length(y)
  w = softmax(params[-(1:n)])
  theta = params[1:n]
  z = S_inv(theta,s,w,grid,z_range)
  return(sum((y-theta)^2/2/s^2 - l_nm(z,s,w,grid)-(z-theta)^2/2/s^2))
}


#'objective function
#'@param theta (theta,w)
#'@param grid prior sds
f_obj_grad = function(params,y,s,grid,z_range,opt_method='L-BFGS-B'){
  n = length(y)
  a = params[-(1:n)]
  w = softmax(a)
  theta = params[1:n]
  z = S_inv(theta,s,w,grid,z_range)
  grad_theta = (z-y)/s^2
  grad_a = -colSums(l_nm_d1_a(z,s,a,grid))
  return(c(grad_theta,c(grad_a)))
}
```

```{r}
ebnm_penalized_inversion = function(x,s,grid,theta_init = NULL,w_init=NULL,z_range=NULL,opt_method = 'L-BFGS-B'){
  n = length(x)
  K = length(grid)
  if(is.null(w_init)){
    w_init = rep(1/K,K)
  }
  if(length(s)==1){
    s = rep(s,n)
  }
  if(is.null(theta_init)){
    theta_init = rep(0,n)
  }
  if(is.null(z_range)){
    z_range = range(x) + c(-1,1)
  }
  params = c(theta_init,w_init)
  out = optim(params,
                   fn=f_obj,
                   gr = f_obj_grad,
                 y=x,
                 s=s,
                 grid=grid,
                 z_range=z_range,
                 method = opt_method)
  
  return(list(posteriorMean = out$par[1:n],w = softmax(out$par[-(1:n)]),opt_res = out))
}
```

```{r}
ploter = function(fit,y,mu,main='estimate prior'){
  plot(y,main=main,col='grey80')
  lines(mu,col='grey60')
  lines(fit$posteriorMean)
  legend('topleft',c('data','true mean','estimated mean'),pch=c(1,NA,NA),lty=c(NA,1,1),col=c('grey80','grey60',1))
}
```

### Init at data $y$.

```{r}
fit = ebnm_penalized_inversion(y,s,grid,theta_init = y,opt_method = 'L-BFGS-B')
plot(grid,fit$w,ylab='w hat')
fit$opt_res$value
```

```{r}
ploter(fit,y,mu,main='estimate prior, init at y')
```


### Init at 0.

```{r}
fit = ebnm_penalized_inversion(y,s,grid,theta_init = rep(0,n),opt_method = 'L-BFGS-B')
plot(grid,fit$w,ylab='w hat')
fit$opt_res$value
```

```{r}
ploter(fit,y,mu,main='estimate prior, init at 0')
```


## ash fit

```{r,warning=FALSE}
fit.ash = ash(y,s,mixcompdist = 'normal',pointmass=FALSE,prior='uniform',mixsd=grid)
plot(y,main='ash fit',col='grey80')
lines(mu,col='grey60')
lines(fit.ash$result$PosteriorMean)
legend('topleft',c('data','true mean','ash posteriorMean'),pch=c(1,NA,NA),lty=c(NA,1,1),col=c('grey80','grey60',1))
```


