---
title: "Overdispersed splitting method dealing with NB data"
author: "DongyueXie"
date: "2022-11-21"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction

We simulate mean parameter $\lambda$ from $\pi_0\delta_0 + \pi_1Exp(0.1)$.

Then generate data using a NB distribution $NB(r,p)$. Then $r(1-p)/p = \lambda$ so $p = r/(r+\lambda)$. The variance is $r(1-p)/p^2 = \lambda + \lambda^2/r$.

What's the corresponding $\sigma^2$ in $Poisson(\exp(\mu+\sigma^2))$?

Since $\exp(\mu+\sigma2/2)=\lambda$, we have $\mu = \log\lambda - \sigma^2/2$. Then by matching the variance of NB and the Poisson model, we solve $(\exp(\sigma^2)-1)\exp(2\mu+\sigma^2) = \lambda^2/r$ and we have $\sigma^2 = \log(1+1/r)$. The smaller the $r$, the larger oversidpersion.

```{r}
library(vebpm)
```

## r = 10

### first run

```{r}
set.seed(12345)
n = 3000
lambda = c(rep(0,n*0.8),rexp(n*0.2,0.1))
r = 10
y = rnbinom(n,r,mu  = lambda)
```

```{r}
sigma2 = log(1+1/r)
sigma2
```

fix $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = log(y+1))
fit$elbo
plot(lambda,col='grey80')
lines(fit$posterior$mean)

fit$fitted_g
```


### second run

```{r}
lambda = c(rep(0,n*0.8),rexp(n*0.2,0.1))
y = rnbinom(n,r,mu  = lambda)
```

fix $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = log(y+1))
fit$elbo
plot(lambda,col='grey80')
lines(fit$posterior$mean)

fit$fitted_g
```


### third run

```{r}
lambda = c(rep(0,n*0.8),rexp(n*0.2,0.1))
y = rnbinom(n,r,mu  = lambda)
```

fix $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = log(y+1))
fit$elbo
plot(lambda,col='grey80')
lines(fit$posterior$mean)

fit$fitted_g
```



## r = 5

### first run

```{r}
set.seed(12345)
n = 3000
lambda = c(rep(0,n*0.8),rexp(n*0.2,0.1))
r = 5
y = rnbinom(n,r,mu  = lambda)
```

```{r}
sigma2 = log(1+1/r)
sigma2
```

fix $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = log(y+1))
fit$elbo
plot(lambda,col='grey80')
lines(fit$posterior$mean)

fit$fitted_g
```


### second run

```{r}
lambda = c(rep(0,n*0.8),rexp(n*0.2,0.1))
y = rnbinom(n,r,mu  = lambda)
```

fix $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = log(y+1))
fit$elbo
plot(lambda,col='grey80')
lines(fit$posterior$mean)

fit$fitted_g
```


### third run

```{r}

lambda = c(rep(0,n*0.8),rexp(n*0.2,0.1))

y = rnbinom(n,r,mu  = lambda)
```

fix $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = log(y+1))
fit$elbo
plot(lambda,col='grey80')
lines(fit$posterior$mean)

fit$fitted_g
```

## r = 50

### first run

```{r}
set.seed(12345)
n = 3000
lambda = c(rep(0,n*0.8),rexp(n*0.2,0.1))
r = 50
y = rnbinom(n,r,mu  = lambda)
```

```{r}
sigma2 = log(1+1/r)
sigma2
```

fix $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = log(y+1))
fit$elbo
plot(lambda,col='grey80')
lines(fit$posterior$mean)

fit$fitted_g
```


### second run

```{r}

lambda = c(rep(0,n*0.8),rexp(n*0.2,0.1))

y = rnbinom(n,r,mu  = lambda)
```

fix $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = log(y+1))
fit$elbo
plot(lambda,col='grey80')
lines(fit$posterior$mean)

fit$fitted_g
```


### third run

```{r}

lambda = c(rep(0,n*0.8),rexp(n*0.2,0.1))

y = rnbinom(n,r,mu  = lambda)
```

fix $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = log(y+1))
fit$elbo
plot(lambda,col='grey80')
lines(fit$posterior$mean)

fit$fitted_g
```


## r = 500

### first run

```{r}
set.seed(12345)
n = 3000
lambda = c(rep(0,n*0.8),rexp(n*0.2,0.1))
r = 500
y = rnbinom(n,r,mu  = lambda)
```

```{r}
sigma2 = log(1+1/r)
sigma2
```

fix $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = log(y+1))
fit$elbo
plot(lambda,col='grey80')
lines(fit$posterior$mean)

fit$fitted_g
```


### second run

```{r}

lambda = c(rep(0,n*0.8),rexp(n*0.2,0.1))

y = rnbinom(n,r,mu  = lambda)
```

fix $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = log(y+1))
fit$elbo
plot(lambda,col='grey80')
lines(fit$posterior$mean)

fit$fitted_g
```


### third run

```{r}

lambda = c(rep(0,n*0.8),rexp(n*0.2,0.1))

y = rnbinom(n,r,mu  = lambda)
```

fix $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = log(y+1))
fit$elbo
plot(lambda,col='grey80')
lines(fit$posterior$mean)

fit$fitted_g
```
