---
title: "Overdispersed splitting method"
author: "DongyueXie"
date: "2022-11-19"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Generate data from Splitting model

We simulate data from the following model,

$$y_i\sim Poisson(\exp(\mu_i)),\mu_i|b_i\sim N(b_i,\sigma^2), b_i\sim \pi_0\delta_0 + \pi_1\ N(0,2).$$
And Run splitting method with fixed and known $\sigma^2$.

```{r}
library(vebpm)
```

## $\sigma^2 = 0.1$

### first run

```{r}
set.seed(12345)
n = 3000
prior_mean = 0
b = c(rep(prior_mean,n*0.8) , rnorm(n*0.2,prior_mean,sqrt(2)))
sigma2 = 0.1
mu = rnorm(n,b,sigma2)
y = rpois(n,exp(mu))
```

fix $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = log(y+1))
fit$elbo
plot(b,col='grey80')
lines(fit$posterior$mean_b)
```

estimate $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = T,mu_pm_init = log(y+1))
fit$elbo
plot(b,col='grey80')
lines(fit$posterior$mean_b)
```

### second run

```{r}
n = 3000
prior_mean = 0
b = c(rep(prior_mean,n*0.8) , rnorm(n*0.2,prior_mean,sqrt(2)))
sigma2 = 0.1
mu = rnorm(n,b,sigma2)
y = rpois(n,exp(mu))
```

fix $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = log(y+1))
fit$elbo
plot(b,col='grey80')
lines(fit$posterior$mean_b)
```

estimate $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = T,mu_pm_init = log(y+1))
fit$elbo
plot(b,col='grey80')
lines(fit$posterior$mean_b)
```


### third run

```{r}
n = 3000
prior_mean = 0
b = c(rep(prior_mean,n*0.8) , rnorm(n*0.2,prior_mean,sqrt(2)))
sigma2 = 0.1
mu = rnorm(n,b,sigma2)
y = rpois(n,exp(mu))
```

fix $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = log(y+1))
fit$elbo
plot(b,col='grey80')
lines(fit$posterior$mean_b)
```

estimate $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = T,mu_pm_init = log(y+1))
fit$elbo
plot(b,col='grey80')
lines(fit$posterior$mean_b)
```


## $\sigma^2 = 0.5$

### first run

```{r}
set.seed(12345)
n = 3000
prior_mean = 0
b = c(rep(prior_mean,n*0.8) , rnorm(n*0.2,prior_mean,sqrt(2)))
sigma2 = 0.5
mu = rnorm(n,b,sigma2)
y = rpois(n,exp(mu))
```

fix $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = log(y+1))
fit$elbo
plot(b,col='grey80')
lines(fit$posterior$mean_b)
```

estimate $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = T,mu_pm_init = log(y+1))
fit$elbo
plot(b,col='grey80')
lines(fit$posterior$mean_b)
```


### second run

```{r}
n = 3000
prior_mean = 0
b = c(rep(prior_mean,n*0.8) , rnorm(n*0.2,prior_mean,sqrt(2)))
sigma2 = 0.5
mu = rnorm(n,b,sigma2)
y = rpois(n,exp(mu))
```

fix $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = log(y+1))
fit$elbo
plot(b,col='grey80')
lines(fit$posterior$mean_b)
```

estimate $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = T,mu_pm_init = log(y+1))
fit$elbo
plot(b,col='grey80')
lines(fit$posterior$mean_b)
```


### third run

```{r}
n = 3000
prior_mean = 0
b = c(rep(prior_mean,n*0.8) , rnorm(n*0.2,prior_mean,sqrt(2)))
sigma2 = 0.5
mu = rnorm(n,b,sigma2)
y = rpois(n,exp(mu))
```

fix $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = log(y+1))
fit$elbo
plot(b,col='grey80')
lines(fit$posterior$mean_b)
```

estimate $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = T,mu_pm_init = log(y+1))
fit$elbo
plot(b,col='grey80')
lines(fit$posterior$mean_b)
```


### set prior mean to 5

```{r}
set.seed(12345)
n = 3000
prior_mean = 5
b = c(rep(prior_mean,n*0.8) , rnorm(n*0.2,prior_mean,sqrt(2)))
sigma2 = 0.5
mu = rnorm(n,b,sigma2)
y = rpois(n,exp(mu))
```

fix $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = log(y+1))
fit$elbo
plot(b,col='grey80')
lines(fit$posterior$mean_b)
```

estimate $\sigma2$

```{r}
fit = pois_mean_split2(y,sigma2=sigma2,est_sigma2 = T,mu_pm_init = log(y+1))
fit$elbo
plot(b,col='grey80')
lines(fit$posterior$mean_b)
```



