---
title: "Overdispersed splitting method dealing with NB data"
author: "DongyueXie"
date: "2022-11-21"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction

We simulate mean parameter $\lambda$ from $\pi_0\delta_0 + \pi_1Exp(0.1)$.

Then generate data using a NB distribution $NB(r,p)$. Then $r(1-p)/p = \lambda$ so $p = r/(r+\lambda)$. The variance is $r(1-p)/p^2 = \lambda + \lambda^2/r$.

What's the corresponding $\sigma^2$ in $Poisson(\exp(\mu+\sigma^2))$?

Since $\exp(\mu+\sigma2/2)=\lambda$, we have $\mu = \log\lambda - \sigma^2/2$. Then by matching the variance of NB and the Poisson model, we solve $(\exp(\sigma^2)-1)\exp(2\mu+\sigma^2) = \lambda^2/r$ and we have $\sigma^2 = \log(1+1/r)$. The smaller the $r$, the larger oversidpersion.

```{r}
library(vebpm)
library(ashr)
```

```{r}
simu_func = function(n_simu=10,n,r,prior_rate=0.1,w=0.8,seed = 12345,n_plot = 3){
  set.seed(seed)
  mse_mean = c()
  mse_non0_mean = c()
  sigma2 = log(1+1/r)
  
  for(i in 1:n_simu){
    lambda = c(rep(0,round(n*w)) , rexp(round(n*(1-w)),prior_rate))
    non0_idx = which(lambda!=0)
    y = rnbinom(n,r,mu  = lambda)
    fit_ash = ash_pois(y)
    fit_split_ash_init = pois_mean_split(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = fit_ash$result$PosteriorMean,tol=1e-3)
    #fit_split_logx = pois_mean_split(y,sigma2=sigma2,est_sigma2 = FALSE,mu_pm_init = NULL)
    # mse_b = rbind(mse_b,c(mse(log(fit_ash$result$PosteriorMean),(b)),
    #                       mse(fit_split_ash_init$posterior$mean_b,b),
    #                       mse(fit_split_logx$posterior$mean_b,b)))
    mse_mean = rbind(mse_mean,c(mse(fit_ash$result$PosteriorMean,lambda),
                                  mse(fit_split_ash_init$posterior$mean_exp_b,lambda),
                                 mse(exp(fit_split_ash_init$posterior$mean_b),lambda)))
    mse_non0_mean = rbind(mse_non0_mean,c(mse(fit_ash$result$PosteriorMean[non0_idx],lambda[non0_idx]),
                              mse(fit_split_ash_init$posterior$mean_exp_b[non0_idx],lambda[non0_idx]),
                              mse(exp(fit_split_ash_init$posterior$mean_b[non0_idx]),lambda[non0_idx])))

    colnames(mse_mean) = c('ash','split','split exp(b_pm)')
    colnames(mse_non0_mean) = c('ash','split','split exp(b_pm)')
    if(i<=n_plot){
      par(mfrow=c(3,1))
      plot(lambda,col='grey80',ylab='lambda',main='ash')
      lines(fit_ash$result$PosteriorMean)
      plot(lambda,col='grey80',ylab='lambda',main='splitting ')
      lines(fit_split_ash_init$posterior$mean_exp_b)
      plot(lambda,col='grey80',ylab='lambda',main='splitting, exp(b_pm)')
      lines(exp(fit_split_ash_init$posterior$mean_b))
    }
    
  }
  return(list(mse_mean = mse_mean,mse_non0_mean = mse_non0_mean,sigma2=sigma2))
  
}
```

## r = 10

```{r,warning=F,fig.width=4,fig.height=6}
res = simu_func(n_simu=10,n=1000,r = 10,n_plot = 10)
colMeans(res$mse_mean)
colMeans(res$mse_non0_mean)
```

## r = 5

```{r,warning=F,fig.width=4,fig.height=6}
res = simu_func(n_simu=10,n=1000,r = 5,n_plot = 10)
```

## r = 1

```{r,warning=F,fig.width=4,fig.height=6}
res = simu_func(n_simu=10,n=1000,r = 1,n_plot = 10)
```

## r = 50

```{r,warning=F,fig.width=4,fig.height=6}
res = simu_func(n_simu=10,n=1000,r = 50,n_plot = 10)
```



