---
title: "Run ebpmf pbmc3k with nonnegative LF while also updating f0"
author: "DongyueXie"
date: "2023-02-09"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Recall the model is 

\[y_{ij}\sim Poisson(l_{i0}f_{j0}\exp(\sum_k l_{ik}f_{jk}+\epsilon_{ij}).\]

I compare the fit between fix $f_{j0} = y_{+j}/\sqrt{y_{++}}$, and update $f_{j0}$.

ALL other parameters are the same.

When estimating $f_{j0}$, I got an error saying:

```{bash,eval=FALSE}
running initial flash fit
Running iterations...
iter 10, elbo=-18401566.4360083, K=8
iter 20, elbo=-17763539.8435548, K=8
iter 30, elbo=-17393989.0758334, K=8
Error in (function (f, p, ..., hessian = FALSE, typsize = rep(1, length(p)),  : 
  missing value in parameter
Calls: ebpmf_log ... parametric_workhorse -> mle_parametric -> do.call -> <Anonymous>
In addition: Warning message:
In handle_standard_errors(x, s) :
  Nonpositive SEs have been replaced by small positive SEs.
Execution halted

```

My general impression is that the point exponential prior implementation in `ebnm` is less stable, and can yield less desirable results, but much faster, [when comparing to unimodal nonnegative prior](example_unimodal_nonnegative_better_than_point_exponential.html).

Since I only got results from first 30 iterations, I'll compare the 30th iterations results.

```{r}
source('code/poisson_STM/plot_factors_general.R')
source('code/poisson_STM/structure_plot.R')
library(fastTopics)
library(Matrix)
library(stm)
data(pbmc_facs)
```

## fix f0

```{r}
fit = readRDS(paste('output/pbmc3k_iteration_results/ebpmf_pbmc3k_nonnegLF_pe_vga3_iter',30,'.rds',sep=''))
plt = plot.factors.general(fit$fit_flash$L.pm,pbmc_facs$samples$subpop,title=paste('pbmc3k nonneg L F iteration',30))
print(plt)
fit$elbo
```

```{r}
structure_plot_general(fit$fit_flash$L.pm,fit$fit_flash$F.pm,pbmc_facs$samples$subpop)
```

```{r}
f00 = colSums(pbmc_facs$counts)/sqrt(sum(pbmc_facs$counts))
```

## update f0

```{r}
fit = readRDS(paste('output/pbmc3k_iteration_results/ebpmf_pbmc3k_nonnegLF_pe_vga3_est_f0_iter30.rds'))
plt = plot.factors.general(fit$fit_flash$L.pm,pbmc_facs$samples$subpop,title=paste('pbmc3k update f0 nonneg L F iteration',30))
print(plt)
fit$elbo
```


```{r}
structure_plot_general(fit$fit_flash$L.pm,fit$fit_flash$F.pm,pbmc_facs$samples$subpop)
```

```{r}
plot(f00,fit$f0,xlab='init f0',ylab='fitted f0',main='fixed vs updated')
abline(a=0,b=1,lty=2)
```






















