---
title: "poisson mean woth log(1+exp(x)) link function"
author: "Dongyue Xie"
date: "2022-05-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Consider the model 

\begin{equation}
\begin{split}
    & y_j\sim \Poi(\log(1+\exp(\mu_j))),
    \\
    &\mu_j\sim g(\cdot).
\end{split}
\end{equation}

Using the lower bound in Seeger and Bouchard(2012), we can perform inference by iteratively fitting EBNM problem.

```{r}
source('code/poisson_mean/pois_mean_log1exp.R')
```

Generate data

```{r}
set.seed(12345)
n = 1000
w = 0.2
lambda = c(rep(0.1,n*(1-w)),rep(10,n*w))
x = rpois(n,lambda)
plot(x,main='',col='grey80',pch=20)
lines(lambda,col='grey50')
legend('topleft',c('data','true mean'),pch=c(20,NA),lty=c(NA,1),col=c('grey80','grey50'))
```


## point-laplace prior, mean = estimate

```{r,warning=FALSE}
out = pois_mean_log1exp(x,maxiter = 300,ebnm_params=list(mode='estimate',prior_family='point_laplace'))
plot(out$obj,type='l',ylab='ELBO',xlab='iteration')

plot(x,main='',col='grey80',pch=20)
lines(log(1+exp(out$m)),col=4)
legend('topleft',c('data','true mean','estimated mean'),pch=c(20,NA,NA),lty=c(NA,1,1),col=c('grey80','grey50',4))
```

```{r}
out$kappa
out$ebnm_res$fitted_g
```

## point-laplace prior, mean = 0

```{r,warning=FALSE}
out = pois_mean_log1exp(x,maxiter = 300,ebnm_params=list(mode=0,prior_family='point_laplace'))
plot(out$obj,type='l',ylab='ELBO',xlab='iteration')

plot(x,main='',col='grey80',pch=20)
lines(log(1+exp(out$m)),col=4)
legend('topleft',c('data','true mean','estimated mean'),pch=c(20,NA,NA),lty=c(NA,1,1),col=c('grey80','grey50',4))
```

```{r}
out$kappa
out$ebnm_res$fitted_g
```

## ash prior, mean = estimate

```{r}
out = pois_mean_log1exp(x,maxiter = 500,ebnm_params=list(mode='estimate',prior_family='normal_scale_mixture'))
plot(out$obj,type='l',ylab='ELBO',xlab='iteration')

plot(x,main='',col='grey80',pch=20)
lines(log(1+exp(out$m)),col=4)
legend('topleft',c('data','true mean','estimated mean'),pch=c(20,NA,NA),lty=c(NA,1,1),col=c('grey80','grey50',4))
```


```{r}
out$ebnm_res$fitted_g
```


## ash prior, mean = 0

```{r}
out = pois_mean_log1exp(x,maxiter = 500,ebnm_params=list(mode=0,prior_family='normal_scale_mixture'))
plot(out$obj,type='l',ylab='ELBO',xlab='iteration')

plot(x,main='',col='grey80',pch=20)
lines(log(1+exp(out$m)),col=4)
legend('topleft',c('data','true mean','estimated mean'),pch=c(20,NA,NA),lty=c(NA,1,1),col=c('grey80','grey50',4))
```


```{r}
out$ebnm_res$fitted_g
```


## Point normal prior, mean = estimate

```{r,warning=FALSE}
out = pois_mean_log1exp(x,maxiter = 100,ebnm_params=list(mode='estimate',prior_family='point_normal'))
plot(out$obj,type='l',ylab='ELBO',xlab='iteration')

plot(x,main='',col='grey80',pch=20)
lines(log(1+exp(out$m)),col=4)
legend('topleft',c('data','true mean','estimated mean'),pch=c(20,NA,NA),lty=c(NA,1,1),col=c('grey80','grey50',4))
```

```{r}
out$kappa
out$ebnm_res$fitted_g
```

Why ELBO is not increasing in this case?

## Point normal prior, mean = 0

```{r,warning=FALSE}
out = pois_mean_log1exp(x,maxiter = 500,ebnm_params=list(mode=0,prior_family='point_normal'))
plot(out$obj,type='l',ylab='ELBO',xlab='iteration')

plot(x,main='',col='grey80',pch=20)
lines(log(1+exp(out$m)),col=4)
legend('topleft',c('data','true mean','estimated mean'),pch=c(20,NA,NA),lty=c(NA,1,1),col=c('grey80','grey50',4))
```

```{r}
out$ebnm_res$fitted_g
```

