---
title: "poisson mean split local optimum"
author: "DongyueXie"
date: "2023-02-06"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

\[y_i\sim Poisson(exp(\mu_j)),\mu_j|b_j\sim N(b_j,\sigma^2),b_j\sim g(\cdot).\]

We show in a simple POisson mean example, there are two local optimums - one where $g_b$ is a point mass, one where $g_b$ is the one we are interested in.

```{r}
library(vebpm)
n = 1000
set.seed(12345)
mu = c(rep(3,100),rep(0,n-100))
x = rpois(n,exp(mu))

plot(x,col='grey80')
lines(exp(mu))

```

In the first example we initialize $\sigma^2=0.5$, ans we can see that the final $sigma^2$ is around 0.22

```{r}
fit0 = pois_mean_split(x,sigma2=0.5)
plot(x,col='grey80')
lines(fit0$posterior$mean_exp_b)
plot(fit0$fitted_g$sigma2_trace)
fit0$fitted_g$g_b
fit0$elbo
```

WE try to initialize $\sigma^2=0.1$, and we can see that the final $sigma^2$ is still around 0.22

```{r}
fit0 = pois_mean_split(x,sigma2=0.1)
plot(x,col='grey80')
lines(fit0$posterior$mean_exp_b)
plot(fit0$fitted_g$sigma2_trace)
fit0$fitted_g$g_b
fit0$elbo
```

So this is a local optimum and is the one that gives the results we'd love to see.

ON the other hand If we initialize $\sigma^2$ to be bigger at $\sigma^2 = 1$, then $\sigma^2$ converges to 1.8 and $g_b$ goes to a point mass.

```{r}
fit0 = pois_mean_split(x,sigma2=1)
plot(x,col='grey80')
lines(fit0$posterior$mean_exp_b)
fit0$fitted_g$sigma2_trace
fit0$fitted_g$g_b
fit0$elbo
```

Try a larger init value of $\sigma^2$ at 3, then same thing happens.

```{r}
fit0 = pois_mean_split(x,sigma2=3)
plot(x,col='grey80')
lines(fit0$posterior$mean_exp_b)
fit0$fitted_g$sigma2_trace
fit0$fitted_g$g_b
fit0$elbo
```


The ELBO is larger when $g_b$ is a point mass and $\sigma^2$ is larger. 

Maybe we should not start with a too large $\sigma^2$ because $g_b$ being a point mass is a local optimum and in most cases we are not very interested in. I find that the starting $\sigma^2$ should not be greater than $var(\bar\mu_i)$, and perhaps smaller.





