---
title: "poisson mean loglik expansion optimization"
author: "Dongyue Xie"
date: "2022-05-26"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction


\[\tilde F(q,g) = l(\bar\mu) + \frac{1}{2}l''(\bar\mu)V_\mu + E_q\log\frac{g(\mu)}{q_\mu}.\]

\[h(\bar \mu,g) = -l(\bar\mu) - \frac{1}{2}\log 2\pi (-l''(\bar\mu))^{-1} +\rho_{g,s^2 = -l''(\bar\mu)^{-1}}(\bar\mu),\]
where 
\[\rho_{g,s^2 = -{l''(\bar\mu)}^{-1}}(\bar\mu) = \min_{q:\E_q\mu=\bar\mu}\left(\frac{1}{2s^2}V_\mu+KL(q||g)+\frac{1}{2}\log2\pi s^2\right).\]

In the Poisson case, the optimization problem is 
\[\min_{\bar\mu,g} h(\bar\mu,g) = e^{\bar\mu}-(y-0.5)\bar\mu +\rho_{g,s^2=\exp(-\bar\mu)}(\bar\mu).\]


We minimize the objective function

\[h(\bar\mu,g) = e^{\bar\mu+e^{-\bar\mu}l'_{nm}(\bar\mu;g,e^{-\bar\mu})}-(y-0.5)(\bar\mu+e^{-\bar\mu}l'_{nm}(\bar\mu;g,e^{-\bar\mu}))\\-l_{nm}(\bar\mu;g,e^{-\bar\mu})-\frac{e^{-\bar\mu}(l'_{nm}(\bar\mu;g,e^{-\bar\mu}))^2}{2}\]


```{r}
#'log marginal likelihood of normal mean model
#'@param x scalar 
#'@param g prior, normal mixture
l_nm = function(x,w,grid){
  s = sqrt(exp(-x))
  log(sum(w*dnorm(x,0,sqrt(grid^2+s^2))))
}

#'first order derivative of log marginal likelihood of normal mean model
#'@param x scalar 
#'@param g prior, normal mixture
l_nm_d1 = function(x,w,grid){
  s = sqrt(exp(-x))
  f = sum(w*dnorm(x,0,sqrt(grid^2+s^2)))
  f_d1 = sum(w*dnorm(x,0,sqrt(grid^2+s^2))*x/(grid^2+s^2))
  f_d1/f
}

softmax = function(a){
  exp(a-max(a))/sum(exp(a-max(a)))
}

#'posterior mean operator
S = function(x,w,grid){
  K = length(w)
  s = sqrt(exp(-x))
  g = normalmix(pi=w,mean=rep(0,K),sd=grid)
  fit.ash = ashr::ash(x,s,g=g,fixg=T)
  fit.ash$result$PosteriorMean
}

#'objective function
#'@param theta (mu_bar,w)
#'@param grid prior sds
f_obj = function(theta,y,grid){
  n = length(y)
  res = 0
  w = softmax(theta[-(1:n)])
  for(i in 1:n){
    res = res + exp(theta[i]+exp(-theta[i])*l_nm_d1(theta[i],w,grid)) - (y[i]-0.5)*(theta[i]+exp(-theta[i])*l_nm_d1(theta[i],w,grid)) - l_nm(theta[i],w,grid) - exp(-theta[i])*(l_nm_d1(theta[i],w,grid))^2/2
  }
  res
}
```

```{r}
set.seed(12345)
n = 100
w0 = 0.9
#lambda = c(rep(0.1,round(n*w0)),rep(10,n-round(n*w0)))
mu = c(rep(0,round(n*w0)),rep(log(20),n-round(n*w0)))
lambda = exp(mu)
y = rpois(n,lambda)
library(ashr)
fit.ash = ashr::ash(log(y+0.01),sqrt(exp(-log(y+0.01))),pointmass=F,mixcompdist='normal',prior='uniform')
#plot(y,col='grey80',xlab='',main='ash')
#lines(exp(fit.ash$result$PosteriorMean),col=4)
grid = fit.ash$fitted_g$sd
K = length(grid)

fit.ash = ashr::ash_pois(y,1,link='log',pointmass=F,mixcompdist='uniform',prior='uniform')
plot(y,col='grey80',main='ash_pois log link')
lines(fit.ash$result$PosteriorMean,col=4)
```


```{r}
mu_init = log(y+1)
w_init = rep(1/K,K)
fit = optim(c(mu_init,w_init),f_obj,method='SANN',y=y,grid=grid)
fit$convergence
fit$value
```

```{r}
plot(mu,col='grey80',xlab='',main='penalized, z',ylim = range(fit$par[1:n]))
lines(fit$par[1:n],col=2)
#plot(log(y+0.01),col='grey80',xlab='',main='penalized, theta = S(z) ')
#lines(lambda,col='grey80')
plot(mu,col='grey80',main='penalized, S(z)',ylim = range(S(fit$par[1:n],softmax(fit$par[-(1:n)]),grid)))
lines(S(fit$par[1:n],softmax(fit$par[-(1:n)]),grid),col=2)

plot(y,col='grey80',xlab='',main='data space',ylim=range(exp(S(fit$par[1:n],softmax(fit$par[-(1:n)]),grid))))
lines(exp(S(fit$par[1:n],softmax(fit$par[-(1:n)]),grid)),col=2)
round(softmax(fit$par[-(1:n)]),2)
plot(grid,softmax(fit$par[-(1:n)]),ylab='w')
```

```{r}
library(nloptr)
fit = nloptr(c(mu_init,w_init),f_obj,
             opts=list(algorithm='NLOPT_LN_NELDERMEAD',maxeval=10000),
             y=y,grid=grid)
z = fit$solution[1:n]
w_est = softmax(fit$solution[-(1:n)])
```

```{r}
plot(mu,col='grey80',xlab='',main='penalized, z',ylim = range(z))
lines(z,col=2)
#plot(log(y+0.01),col='grey80',xlab='',main='penalized, theta = S(z) ')
#lines(lambda,col='grey80')
plot(mu,col='grey80',main='penalized, S(z)',ylim = range(S(z,w_est,grid)))
lines(S(z,w_est,grid),col=2)

plot(y,col='grey80',xlab='',main='data space',ylim=range(exp(S(z,w_est,grid))))
lines(exp(S(z,w_est,grid)),col=2)
round(w_est,2)
plot(grid,w_est,ylab='w')
```
