---
title: "poisson mean via penalized form"
author: "Dongyue Xie"
date: "2022-05-16"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r}
source("code/poisson_mean/pois_mean_penalized.R")
```

We solve the following Poisson mean problem by optimization method, \[x_i\sim Poisson(\exp\mu_i),\mu_i\sim g(\cdot)\]

We first generate data and fit poisson ash with log link.
```{r,warning=FALSE}
library(ashr)
set.seed(12345)
w = c(0.8,0.2)
n = 100
mu = c(rnorm(n*w[1],0,0.01),rnorm(n*w[2],0,2))
x = rpois(n,exp(mu))

fit.ash = ash(x,1,lik=lik_pois(x,link='log'),mode=0)
fit.ash$fitted_g
plot(x,col='grey50',main='ash fit',ylim=c(0,150))
lines(fit.ash$result$PosteriorMean,col=4)
legend('topleft',c('data','ash posterior mean'), pch=c(1,NA),lty=c(NA,1),col=c('grey50',4))

plot(mu,col='grey50',main='ash fit',ylim=c(-2,6))
lines(log(fit.ash$result$PosteriorMean))
legend('topleft',c('true mu','log(ash posterior mean)'), pch=c(1,NA),lty=c(NA,1),col=c('grey50',4))
```

## known prior

Now we fit the penalize poisson mean problem with known prior.

Use Nelder-Mead:

```{r}
grid=c(0.01,2)
fit = pois_mean_penalized_optim(x,w=w,grid=grid,est_w=FALSE,z_init = log(1+x),opt_method = 'Nelder-Mead')
plot(L_grad_known_g(fit$fit$par,x,w,grid),ylab='gradient',main='check gradient')

plot(fit$z,col='grey50')
lines(log(1+x),type='p',pch=8,col='grey50')
lines(mu,col='grey80',type='p',pch=20)
lines(fit$m,col=4)
legend('topleft',c('log(1+x)','z','true mu','mu hat'), pch=c(8,1,20,NA),lty=c(NA,NA,NA,1),col=c('grey50','grey50','grey80',4))

fit$s2

L_known_g(fit$fit$par,x,w,grid)
```

Use BFGS:
```{r}
fit = pois_mean_penalized_optim(x,w=w,grid=grid,est_w=FALSE,z_init = log(1+x),opt_method = 'BFGS')
plot(L_grad_known_g(fit$fit$par,x,w,grid),ylab='gradient',main='check gradient')

plot(fit$z,col='grey50')
lines(log(1+x),type='p',pch=8,col='grey50')
lines(mu,col='grey80',type='p',pch=20)
lines(fit$m,col=4)
legend('topleft',c('log(1+x)','z','true mu','mu hat'), pch=c(8,1,20,NA),lty=c(NA,NA,NA,1),col=c('grey50','grey50','grey80',4))

fit$s2

L_known_g(fit$fit$par,x,w,grid)
```


Use CG:
```{r}
fit = pois_mean_penalized_optim(x,w=w,grid=grid,est_w=FALSE,z_init = log(1+x),opt_method = 'CG')
plot(L_grad_known_g(fit$fit$par,x,w,grid),ylab='gradient',main='check gradient')

plot(fit$z,col='grey50')
lines(log(1+x),type='p',pch=8,col='grey50')
lines(mu,col='grey80',type='p',pch=20)
lines(fit$m,col=4)
legend('topleft',c('log(1+x)','z','true mu','mu hat'), pch=c(8,1,20,NA),lty=c(NA,NA,NA,1),col=c('grey50','grey50','grey80',4))

fit$s2

L_known_g(fit$fit$par,x,w,grid)
```


Use `nleqslv` for solving gradient = 0 directly.

```{r}
library(nleqslv)
grid=c(0.01,2)
fit = pois_mean_penalized_nleqslv(x,w=w,grid=grid,est_w=FALSE,z_init = log(1+x),opt_method = 'Broyden')
plot(L_grad_known_g(fit$fit$x,x,w,grid),ylab='gradient',main='check gradient')

plot(fit$z,col='grey50')
lines(log(1+x),type='p',pch=8,col='grey50')
lines(mu,col='grey80',type='p',pch=20)
lines(fit$m,col=4)
legend('topleft',c('log(1+x)','z','true mu','mu hat'), pch=c(8,1,20,NA),lty=c(NA,NA,NA,1),col=c('grey50','grey50','grey80',4))

fit$s2

L_known_g(fit$fit$x,x,w,grid)
```


```{r}
fit = pois_mean_penalized_nleqslv(x,w=w,grid=grid,est_w=FALSE,z_init = log(1+x),opt_method = 'Newton')
plot(L_grad_known_g(fit$fit$x,x,w,grid),ylab='gradient',main='check gradient')

plot(fit$z,col='grey50')
lines(log(1+x),type='p',pch=8,col='grey50')
lines(mu,col='grey80',type='p',pch=20)
lines(fit$m,col=4)
legend('topleft',c('log(1+x)','z','true mu','mu hat'), pch=c(8,1,20,NA),lty=c(NA,NA,NA,1),col=c('grey50','grey50','grey80',4))

fit$s2

L_known_g(fit$fit$x,x,w,grid)
```


## estimating prior

```{r}
fit.ash = ashr::ash(log(x+0.01),sqrt(exp(-log(x+0.01))),pointmass=F,mixcompdist='normal',prior='uniform')
grid = fit.ash$fitted_g$sd
K = length(grid)
```

opt_method = 'BFGS'

```{r}
fit = pois_mean_penalized_optim(x,w=NULL,grid=grid,est_w=TRUE,z_init = log(1+x),opt_method = 'BFGS')
```

```{r}
round(fit$w,3)
plot(grid,fit$w)
```

```{r}
plot(L_grad(fit$fit$par,x,grid),ylab='gradient',main='check gradient')

plot(fit$z,col='grey50')
lines(log(1+x),type='p',pch=8,col='grey50')
lines(mu,col='grey80',type='p',pch=20)
lines(fit$m,col=4)
legend('topleft',c('log(1+x)','z','true mu','mu hat'), pch=c(8,1,20,NA),lty=c(NA,NA,NA,1),col=c('grey50','grey50','grey80',4))

fit$s2
```

opt_method = 'Nelder-Mead'

```{r}
fit = pois_mean_penalized_optim(x,w=NULL,grid=grid,est_w=TRUE,z_init = log(1+x),opt_method = 'Nelder-Mead')
```

```{r}
round(fit$w,3)
plot(grid,fit$w)
```

```{r}
plot(L_grad(fit$fit$par,x,grid),ylab='gradient',main='check gradient')

plot(fit$z,col='grey50')
lines(log(1+x),type='p',pch=8,col='grey50')
lines(mu,col='grey80',type='p',pch=20)
lines(fit$m,col=4)
legend('topleft',c('log(1+x)','z','true mu','mu hat'), pch=c(8,1,20,NA),lty=c(NA,NA,NA,1),col=c('grey50','grey50','grey80',4))

fit$s2
```


opt_method = 'L-BFGS-B'

```{r}
fit = pois_mean_penalized_optim(x,w=NULL,grid=grid,est_w=TRUE,z_init = log(1+x),opt_method = 'L-BFGS-B')
```

```{r}
round(fit$w,3)
plot(grid,fit$w)
```

```{r}
plot(L_grad(fit$fit$par,x,grid),ylab='gradient',main='check gradient')

plot(fit$z,col='grey50')
lines(log(1+x),type='p',pch=8,col='grey50')
lines(mu,col='grey80',type='p',pch=20)
lines(fit$m,col=4)
legend('topleft',c('log(1+x)','z','true mu','mu hat'), pch=c(8,1,20,NA),lty=c(NA,NA,NA,1),col=c('grey50','grey50','grey80',4))

fit$s2
```


opt_method = 'CG'

```{r}
fit = pois_mean_penalized_optim(x,w=NULL,grid=grid,est_w=TRUE,z_init = log(1+x),opt_method = 'CG')
```

```{r}
round(fit$w,3)
plot(grid,fit$w)
```

```{r}
plot(L_grad(fit$fit$par,x,grid),ylab='gradient',main='check gradient')

plot(fit$z,col='grey50')
lines(log(1+x),type='p',pch=8,col='grey50')
lines(mu,col='grey80',type='p',pch=20)
lines(fit$m,col=4)
legend('topleft',c('log(1+x)','z','true mu','mu hat'), pch=c(8,1,20,NA),lty=c(NA,NA,NA,1),col=c('grey50','grey50','grey80',4))

fit$s2
```


Use `nleqslv` for solving gradient = 0 directly.


```{r}
fit = pois_mean_penalized_nleqslv(x,w=NULL,grid=grid,est_w=TRUE,z_init = log(1+x),opt_method = 'Broyden')
plot(L_grad(fit$fit$x,x,grid),ylab='gradient',main='check gradient')

plot(fit$z,col='grey50')
lines(log(1+x),type='p',pch=8,col='grey50')
lines(mu,col='grey80',type='p',pch=20)
lines(fit$m,col=4)
legend('topleft',c('log(1+x)','z','true mu','mu hat'), pch=c(8,1,20,NA),lty=c(NA,NA,NA,1),col=c('grey50','grey50','grey80',4))

round(fit$w,3)

plot(grid,fit$w)

fit$s2
```

```{r}
fit = pois_mean_penalized_nleqslv(x,w=NULL,grid=grid,est_w=TRUE,z_init = log(1+x),opt_method = 'Newton')
plot(L_grad(fit$fit$x,x,grid),ylab='gradient',main='check gradient')

plot(fit$z,col='grey50')
lines(log(1+x),type='p',pch=8,col='grey50')
lines(mu,col='grey80',type='p',pch=20)
lines(fit$m,col=4)
legend('topleft',c('log(1+x)','z','true mu','mu hat'), pch=c(8,1,20,NA),lty=c(NA,NA,NA,1),col=c('grey50','grey50','grey80',4))

round(fit$w,3)

plot(grid,fit$w)

fit$s2
```




