---
title: "Poisson mean with log(1+exp(x)) link function"
author: "Dongyue Xie"
date: "2022-05-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Consider the model 

\begin{equation}
\begin{split}
    & y_j\sim Poisson(\log(1+\exp(\mu_j))),
    \\
    &\mu_j\sim g(\cdot).
\end{split}
\end{equation}

Using the lower bound in Seeger and Bouchard(2012), we can perform inference by iteratively fitting EBNM problem.

```{r}
source('code/poisson_mean/pois_mean_log1exp.R')
```

```{r}
rmse = function(x,y){sqrt(mean((x-y)^2))}
```

```{r}
set.seed(12345)
n = 1000
w = 0.2
mu = c(rep(6,n*0.9),rnorm(n*0.1,6,2))
lambda = log(1+exp(mu))
x = rpois(n,lambda)
plot(x,main='',col='grey80',pch=20)
lines(lambda,col='grey50')
legend('topleft',c('data','true mean'),pch=c(20,NA),lty=c(NA,1),col=c('grey80','grey50'))
```

We initialize $\tilde\mu$ at $x$. (init from a contant won't work)

```{r,warning=FALSE}
out = pois_mean_log1exp(x,maxiter = 1000,ebnm_params=list(mode='estimate',prior_family='point_laplace'))
plot(out$obj,type='l',ylab='ELBO',xlab='iteration')
```


```{r}
plot(x,main='data space',col='grey80',pch=20)
lines(lambda,col='grey50')
lines(log(1+exp(out$posteriorMean)),col=4)
legend('topleft',c('data','true mean','estimated mean'),pch=c(20,NA,NA),lty=c(NA,1,1),col=c('grey80','grey50',4))
```

```{r}
plot(log(exp(x)-1),main='latent space',col='grey80',pch=20,ylim = range(c(x,mu,out$posteriorMean)))
lines(log(exp(lambda)-1),col='grey50')
lines(out$posteriorMean,col=4)
legend('topleft',c('data','true latent','estimated latent'),pch=c(20,NA,NA),lty=c(NA,1,1),col=c('grey80','grey50',4))
```

```{r}
out$kappa
out$ebnm_res$fitted_g
```

```{r}
rmse(out$posteriorMean,mu)
```

We now change the largest observation so that it changes the $s$ in EBNM problem.

```{r}
max_x = c(20,30,40,50,60,70,80,90,100,200)

n_iters = c()
errors = c()

for(ll in max_x){
  kapa = 0.25+0.17*ll
  out = pois_mean_log1exp(x,maxiter = 10000,ebnm_params=list(mode='estimate',prior_family='point_laplace'),kapa=kapa)
  n_iters = c(n_iters,length(out$obj_value))
  errors = c(errors,rmse(mu,out$posteriorMean))
  plot(log(exp(x)-1),main='latent space',col='grey80',pch=20,ylim = range(c(x,mu,out$posteriorMean)))
  lines(log(exp(lambda)-1),col='grey50')
  lines(out$posteriorMean,col=4)
  legend('topleft',c('data','true latent','estimated latent'),pch=c(20,NA,NA),lty=c(NA,1,1),col=c('grey80','grey50',4))
}
```

```{r}
plot(max_x,n_iters,type='b')
plot(max_x,errors,type='b')
```

