---
title: "poisson smooth split local optimum"
author: "DongyueXie"
date: "2023-02-07"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

\[y_i\sim Poisson(exp(\mu_j)),\mu_j|b_j\sim N(b_j,\sigma^2),b\sim g(\cdot).\]

where $g(\cdot)$ is a wavelet prior.

## A simple block function

```{r}
library(smashrgen)
library(wavethresh)

set.seed(12345)
n=2^9
sigma=0
mu=c(rep(0.3,n/4), rep(3, n/4), rep(10, n/4), rep(0.3, n/4))
x = rpois(n,exp(log(mu)+rnorm(n,sd=sigma)))
fit = pois_smooth_split(x,maxiter=300,verbose = T,tol=1e-10)
```

```{r}
plot(x,col='grey80')
lines(mu,col='grey60')
lines(fit$posterior$mean_smooth)
```

we see that the $\sigma^2$ converges to 0.

```{r}
plot(fit$fitted_g$sigma2_trace,type='l')
```

And the priors on wavelet coefficients converge to a point mass. At scale 0 and 1, they are pointmass at non-zero posoitons, and at other scales, they are point mass at 0.

```{r}
fit$fitted_g
```

The ELBO is 
```{r}
fit$elbo
```

Let's try to initialize at a different value and let it get stuck at a local optimum. In this case, 

```{r}
 fit = pois_smooth_split(x,maxiter=300,verbose = T,tol=1e-10,m_init=rep(mean(x),n),sigma2_init = 0.01)
```

we see that the $\sigma^2$ converges to 2.

```{r}
plot(fit$fitted_g$sigma2_trace,type='l')
```

And the priors on wavelet coefficients converge to a point mass at 0. 

```{r}
fit$fitted_g
```

The ELBO is 
```{r}
fit$elbo
```


## A simple curve

We use non-haar wavelet here. THe function is spike.

```{r}
set.seed(12345)
n=2^9
count_size = 20
sigma=0.5
t = seq(0,1,length.out = n)
b = smashrgen:::spike.f(t)
b = b/(max(b)/count_size)
x = rpois(n,exp(log(b)+rnorm(n,sd=sigma)))
plot(x,col='grey80')
lines(b,col='grey50')
```
```{r}
fit = pois_smooth_split(x,maxiter=300,verbose = T,tol=1e-10,filter.number = 8)
```

```{r}
plot(x,col='grey80')
lines(b,col='grey60')
lines(fit$posterior$mean_smooth,col=4)
```

The $\sigma^2$ converges to 0.4.

```{r}
plot(fit$fitted_g$sigma2_trace)
fit$fitted_g$sigma2
```

And the priors on wavelet coefficients converge to a point mass at finer level. 

At scale 5 and 2, the prior converges to a mixture of two normals.

At scale 0 and 1, they are point-mass at non-zero positions.

```{r}
fit$fitted_g
```

The ELBO is 
```{r}
fit$elbo
```

Let's try to initialize at a different value and let it get stuck at a local optimum. In this case, 

```{r}
 fit = pois_smooth_split(x,maxiter=300,verbose = T,tol=1e-10,m_init=rep(mean(x),n),sigma2_init = 0.01)
```
```{r}
plot(x,col='grey80')
lines(b,col='grey60')
lines(fit$posterior$mean_smooth,col=4)
```

The $\sigma^2$ converges to 5.47.

```{r}
plot(fit$fitted_g$sigma2_trace)
fit$fitted_g$sigma2
```

And the priors on wavelet coefficients converge to a point mass at 0.

```{r}
fit$fitted_g
```

The ELBO is 
```{r}
fit$elbo
```

What if I still use a Haar wavelet?

```{r}
fit = pois_smooth_split(x,maxiter=300,verbose = T,tol=1e-10,filter.number = 1)
```


```{r}
plot(x,col='grey80')
lines(b,col='grey60')
lines(fit$posterior$mean_smooth,col=4)
```

The $\sigma^2$ converges to 0.8.

```{r}
plot(fit$fitted_g$sigma2_trace)
fit$fitted_g$sigma2
```

The prior does not converge to point-mass at scale 4,5,6.

The ELBO is 
```{r}
fit$elbo
```

## Observations

It seems that for the smoothing problem:

If $g$ goes to pointmass at 0, then estimated $\sigma^2$ is large. This is the same as in Poisson mean problem.

If $g$ goes to non-point mass(In wavelet case, some of the them maybe point mass, but at least one of them is not), then $\sigma^2$ can converge to smaller value, but not 0.

If $g$ goes to pointmass's, but not all at 0. This corresponding to the first example. Then in this case $\sigma^2$ seems to be able to converge to 0. And apparently the estimated curve is close to the true one, as shown in plot.



