---
title: "run PMF on pbmc data, 3 cells, nonnegative Loadings, more iterations"
author: "DongyueXie"
date: "2022-12-18"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

In a [previously run](run_PMF_on_pbmc_3cells_nonnegativeL.html), there were 200 iterations but the plot of ELBO shows that the algorithm does not converge.

Here I ran it 1000 iterations.

```{r}
rmse = function(x,y){return(sqrt(mean((x-y)^2)))}
res = readRDS('/project2/mstephens/dongyue/poisson_mf/pbmc_3cells_Sij_nonnegativeL.rds')
fit = res$fit
```

```{r}
fit$run_time
plot(fit$eblo_trace,type='l')
```




```{r}
fit$fit_flash$n.factors
plot(fit$K_trace)
plot(fit$sigma2,ylab = 'sigma2',xlab='gene',col='grey50')
```

Use Jason's method for visualizing loadings:

```{r}
library(fastTopics)
library(stm)
library(Matrix)
data(pbmc_facs)

### use three cells
cells = pbmc_facs$samples$subpop%in%c('B cell', 'NK cell','CD34+')
cell_names =pbmc_facs$samples$subpop[cells]
```

```{r}
source('code/poisson_STM/plot_factors.R')
```

```{r}
plot.factors(fit$fit_flash,cell_names,title='splitting PMF')
```

## RUn time

RUn time analysis

```{r}
unlist(lapply(fit$run_time_break_down,median))
```

Each iteration takes about 13 seconds. 

The most time consuming steps are vga, greedy, and backfitting.

THe vga solves n*p optimization problem using newton's method. 

THe add_greedy command needs to perform a rank-1 init every time, while from the plot of K_tace, we saw the K is stabilized after a few iterations. So we may skip the add greedy step after certain iterations, or do add greedy less often as more and more iterations are done.

The convergence takes more than 1000 iterations. Do we really need such many iterations? How can we accelarete the convergence?



