---
title: "run PMF on pbmc data"
author: "DongyueXie"
date: "2022-12-06"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

I take the pbmc data from `fastTopics` package, and run splitting PMF on the dataset.

```{r}
library(fastTopics)
library(Matrix)
library(stm)
data(pbmc_facs)
counts <- pbmc_facs$counts
table(pbmc_facs$samples$subpop)
hist(rowSums(counts),breaks = 100,main='library size')
sum(counts != 0)/prod(dim(counts))
hist(colMeans(counts!=0),breaks = 100,main='gene expression sparsity')
```

There are 5 main cell types and 16791 genes.

I set the scaling factors as $s_{ij} = \frac{y_{i+}y_{+j}}{y_{++}}$. For comparison, I also fit `flash` on transformed count data, as $\tilde{y}_{ij} = \log(1+\frac{y_{ij}}{s_{ij}}\frac{a_j}{0.5})$ where $a_j = median(s_{\cdot j})$. This transformation is derived from $\tilde{y}_{ij} = \log(\frac{y_{ij}}{s_{ij}}+\frac{0.5}{a_j})$. PMF splitting found 8 topics while  `flash` was not able to terminate at $Kmax = 50$.

```{r}
fit = readRDS('output/poisson_MF_simulation/pbmc.rds')
fit_flashier = readRDS('output/poisson_MF_simulation/fit_flashier_pbmc.rds')
```


```{r}
fit$run_time
length(fit$eblo_trace)
```

```{r}
fit$fit_flash$n.factors
plot(fit$K_trace)
plot(fit$sigma2,ylab = 'sigma2',xlab='gene',col='grey50')
fit$fit_flash$pve
```

```{r}
plot(colSums(counts/c(rowSums(counts)))/dim(counts)[1],fit$sigma2,xlab='gene mean count(after library size adjustment)')
plot(colSums(counts==0)/dim(counts)[1],fit$sigma2,xlab='sparsity')
```

There's trend that the sparser the gene , the larger $\sigma^2$.


```{r,eval=F}
cell_names = as.character(pbmc_facs$samples$subpop)
colors = 1:5
cell_types = unique(cell_names)
color_cell = colors[match(cell_names,cell_types)]
par(mfrow=c(1,1))
plot(fit$fit_flash$L.pm[,1],fit$fit_flash$L.pm[,2],col=color_cell,xlab='third loading',ylab='first loading')
legend(c('topright'),cell_types,col=1:5,pch=c(1,1,1,1,1))
```


Use Jason's method for visualizing loadings:

```{r}
source('code/poisson_STM/plot_factors.R')
```

```{r}
cell_names = as.character(pbmc_facs$samples$subpop)
plot.factors(fit$fit_flash,cell_names,title='splitting PMF')
```

```{r}
plot.factors(fit_flashier,cell_names,kset = c(1:8),title='flashier')
plot.factors(fit_flashier,cell_names,kset = c(1:30),title='flashier')
```

Some of the loadings from flashier such as 3 to 6 are local. So I filtered out genes that with nonzero counts in fewer than 10 cells, and refit flash.

```{r}
fit_flashier = readRDS('output/poisson_MF_simulation/fit_flashier_pbmc_filter_gene.rds')
plot.factors(fit_flashier,cell_names,kset = c(1:8),title='flashier')
plot.factors(fit_flashier,cell_names,kset = c(1:30),title='flashier')
```


## Plot of factors

```{r}
for(k in 1:fit$fit_flash$n.factors){
  plot(fit$fit_flash$F.pm[,k],xlab='gene',ylab='',main=paste('factor',k))
}
```

