---
title: "sigma2 in splitting method"
author: "DongyueXie"
date: "2022-11-16"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

In some cases when we see that [splitting and GG are very similar in terms of MSE.](exp_prior_runtime.html).

```{r}
out = readRDS('output/poisson_mean_simulation/poisson_mean/point_exp_prior32_n_1000_mean10.rds')
```

Why?

```{r}
out$output[[1]]$fitted_model$GG$fitted_g
```

```{r}
out$output[[1]]$fitted_model$split$fitted_g
```

```{r}
out$output[[2]]$fitted_model$GG$fitted_g
```

```{r}
out$output[[2]]$fitted_model$split$fitted_g
```

Because the fitted priors are the same!

In the splitting method  we let 

$$y\sim p(\mu),\mu|b\sim N(b,\sigma^2),b\sim g(\cdot)$$

There are two extreme cases: 1. if $\sigma2=0$ then this is equivalent to $\mu\sim g()$; 2. if $g(\cdot) = \delta_\beta$, then this is equivalent to $\mu\sim N(\beta,\sigma^2)$.



What if we fix $sigma2$ at a small value?

```{r}
library(vebpm)
fit = pois_mean_split(out$sim_data$X[1,],sigma2=0.01,est_sigma2 = FALSE)
```

```{r}
fit$fitted_g
```

```{r}
plot(out$sim_data$X[1,],col='grey80')
lines(fit$posterior$mean)
```

```{r}
plot(out$sim_data$Mean[1,],col='grey80')
lines(fit$posterior$mean)
```

This is due to I initialize $\bar b$ at 0 and $\sigma^2$ is small so $\bar\mu$ would be small, unless for those 5 large values as seen in the plot.

How about changing the init for $\bar b$?

Let's init it at true values. But true log_mean have $-Inf$ values...

```{r,warning=FALSE}
temp = out$sim_data$log_Mean[1,]
temp[temp==-Inf] = -10
fit = pois_mean_split(out$sim_data$X[1,],sigma2=0.01,est_sigma2 = FALSE,b_pm_init = temp,ebnm_params = list(prior_family='normal_scale_mixture'))
```

```{r}
fit$fitted_g
```

```{r}
plot(out$sim_data$X[1,],col='grey80')
lines(fit$posterior$mean)
```

```{r}
plot(out$sim_data$Mean[1,],col='grey80')
lines(fit$posterior$mean)
```


Initialize at ash output:
```{r,warning=FALSE}
fit = pois_mean_split(out$sim_data$X[1,],sigma2=0.01,est_sigma2 = FALSE,b_pm_init =out$output[[1]]$fitted_model$ash_pois_identity$posterior$mean_log,ebnm_params = list(prior_family='normal_scale_mixture'))
```

```{r}
fit$fitted_g
```

```{r}
plot(out$sim_data$X[1,],col='grey80')
lines(fit$posterior$mean)
```

```{r}
plot(out$sim_data$Mean[1,],col='grey80')
lines(fit$posterior$mean)
```
