---
title: "simulation pbmc fasttpoics nonnegtive Loading and Factors"
author: "DongyueXie"
date: "2023-02-22"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

This is a simulation comparing ebpmf and flash on factorizing Poisson matrix. 

To make the simulated dataset close to a real single cell data, I fitted an ebpmf model with nonnegtive loadings and factors on a PBMC single cell data from `fastTopics` package. [See fitted results here.](pbmc3k_infasttopic_result.html)

Then I generated data from the fitted model, and repeated 3 times. When simulating data, The factors and loadings are chosen to be the posterior mean of the fitted model. The `flash` model was fit on transformed count data, as $\tilde{y}_{ij} = \log(1+\frac{a\times y_{ij}}{0.5\times s_{ij}})$ where $a = median(s_i), s_i = \sum_jy_{ij}$. This transformation is derived from $\tilde{y}_{ij} = \log(\frac{y_{ij}}{s_{i}}+\frac{0.5}{a}) = \log (1+\frac{a\times y_{ij}}{0.5\times s_i}) + \log \frac{0.5}{a}$. 


The true dimension is 7+2. **Maybe I should have selected top 5+2 for simulation!**

```{r}
fit = readRDS("~/Documents/myproj/gsmash/output/pbmc_fasttopics/pbmc3k_nonnegLF_pe.rds")
fit$fit_flash$pve
```

```{r}
rmse = function(x,y){return(sqrt(mean((x-y)^2)))}
res = readRDS('output/ebpmf_simulation/pbmc_fasttopics_simu.rds')
```

```{r}
K_hat = c()
for(i in 1:length(res$output)){
  K_hat = rbind(K_hat,c(res$output[[i]]$fitted_model$flash$n.factors,res$output[[i]]$fitted_model$ebpmf$fit_flash$n.factors))
}
colnames(K_hat) = c('flash','ebpmf')
K_hat
```


```{r}
library(fastTopics)
library(Matrix)
library(gridExtra)
source('code/poisson_STM/plot_factors.R')
source('code/poisson_STM/plot_factors_general.R')
source('code/poisson_STM/structure_plot.R')
data("pbmc_facs")
cell_names = pbmc_facs$samples$subpop
```


```{r}
plot0=plot.factors.general(res$sim_data$Loading,res$sim_data$Factor,cell.types=cell_names,title='True Loadings',nonnegative = T)
for(i in 1:length(res$output)){
  plot1 = plot.factors(res$output[[i]]$fitted_model$flash,cell.types=cell_names,title='flash',nonnegative = T)
  plot2 = plot.factors(res$output[[i]]$fitted_model$ebpmf$fit_flash,cell.types=cell_names,title='ebpmf',nonnegative = T)
  grid.arrange(plot1, plot0,plot2, ncol=3)
}
```


```{r}
plot0=structure_plot_general(res$sim_data$Loading,res$sim_data$Factor,grouping =cell_names,title = 'True',print_plot = F) 
for(i in 1:length(res$output)){
  plot1 = structure_plot_general(res$output[[i]]$fitted_model$flash$L.pm,
                                 res$output[[i]]$fitted_model$flash$F.pm
                                 ,grouping=cell_names,
                                 title='flash',
                                 print_plot = F)
  plot2 = structure_plot_general(res$output[[i]]$fitted_model$ebpmf$fit_flash$L.pm,
                                 res$output[[i]]$fitted_model$ebpmf$fit_flash$F.pm,
                                 grouping =cell_names,
                                 title='ebpmf',
                                 print_plot = F)
  grid.arrange(plot1, plot0,plot2, nrow=3)
}
```
