---
title: "simulation study of initial variance influence on model fitting"
author: "DongyueXie"
date: "2023-02-24"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

So in a previous study, I found that [different initial variance can lead to different structures](pbmc_infasttopics_init_vga_effect.html).


In this study, I ran the following simulation:

A. Use the fitted model that inits with small variance to generate data. Specifically, the 1e-6, 60th iter model.

B. Use the fitted model that inits with larger variance to generate data. Specifically, the 1e-2, 60th iter model.

For both A,B, I tried to fit ebpmf model with small and large init variances. Each experiment is repeated 5 times.


```{r}
library(fastTopics)
library(Matrix)
library(gridExtra)
source('code/poisson_STM/plot_factors.R')
source('code/poisson_STM/plot_factors_general.R')
source('code/poisson_STM/structure_plot.R')
source('code/poisson_STM/get_loadings_order.R')
data("pbmc_facs")
cell_names = pbmc_facs$samples$subpop
```


## Small true variance, small init variance

```{r}
res = readRDS('/project2/mstephens/dongyue/poisson_mf/pbmc3k_simulation/simu_pbmc_fasttopics_small_true_var_small_init_var.rds')
K_hat = c()
for(i in 1:length(res$output)){
  K_hat = rbind(K_hat,c(res$output[[i]]$fitted_model$flash$n.factors,res$output[[i]]$fitted_model$ebpmf$fit_flash$n.factors))
}
colnames(K_hat) = c('flash','ebpmf')
K_hat
```

```{r,fig.width=16,fig.height=4}
plot0=plot.factors.general(res$sim_data$Loading,
                           cell.types=cell_names,title='True Loadings',
                           nonnegative = T,kset = 3:8)
for(i in 1:length(res$output)){
  plot1 = plot.factors(res$output[[i]]$fitted_model$flash,
                       cell.types=cell_names,title='flash',
                       nonnegative = T)
  plot2 = plot.factors(res$output[[i]]$fitted_model$ebpmf$fit_flash,
                       cell.types=cell_names,title='ebpmf',
                       nonnegative = T,rm_kset = c(1,2))
  grid.arrange(plot1, plot0,plot2, ncol=3)
}
```

```{r}
loadings_order = get_loadings_order(res$sim_data$Loading,res$sim_data$Factor,grouping = cell_names,n_samples = 5000)
plot0=structure_plot_general(res$sim_data$Loading,res$sim_data$Factor,grouping =cell_names,title = 'True',print_plot = F,loadings_order=loadings_order) 
for(i in 1:length(res$output)){
  plot1 = structure_plot_general(res$output[[i]]$fitted_model$flash$L.pm,
                                 res$output[[i]]$fitted_model$flash$F.pm
                                 ,grouping=cell_names,
                                 title='flash',
                                 print_plot = F,
                                 loadings_order=loadings_order,
                                 remove_l0f0 = F)
  plot2 = structure_plot_general(res$output[[i]]$fitted_model$ebpmf$fit_flash$L.pm,
                                 res$output[[i]]$fitted_model$ebpmf$fit_flash$F.pm,
                                 grouping =cell_names,
                                 title='ebpmf',
                                 print_plot = F,
                                 loadings_order=loadings_order)
  grid.arrange(plot1, plot0,plot2, nrow=3)
}
```

```{r}
fit = readRDS('/project2/mstephens/dongyue/poisson_mf/pbmc_fasttopics/init_tol_effect/nonnegLF_pe_inittol1e6_iter60.rds')
for(i in 1:length(res$output)){
  plot(fit$sigma2,res$output[[i]]$fitted_model$ebpmf$sigma2,col='grey80',pch=20,xlab='true',ylab='estimated',main='variance estimation')
  abline(a=0,b=1)
}

```

## Small true variance, large init variance


```{r}
res = readRDS('/project2/mstephens/dongyue/poisson_mf/pbmc3k_simulation/simu_pbmc_fasttopics_small_true_var_large_init_var.rds')
K_hat = c()
for(i in 1:4){
  K_hat = rbind(K_hat,c(res$output[[i]]$fitted_model$flash$n.factors,res$output[[i]]$fitted_model$ebpmf$fit_flash$n.factors))
}
colnames(K_hat) = c('flash','ebpmf')
K_hat
```

```{r}
loadings_order = get_loadings_order(res$sim_data$Loading,res$sim_data$Factor,grouping = cell_names,n_samples = 5000)
plot0=structure_plot_general(res$sim_data$Loading,res$sim_data$Factor,grouping =cell_names,title = 'True',print_plot = F,loadings_order = loadings_order) 
for(i in 1:4){
  plot1 = structure_plot_general(res$output[[i]]$fitted_model$flash$L.pm,
                                 res$output[[i]]$fitted_model$flash$F.pm
                                 ,grouping=cell_names,
                                 title='flash',
                                 print_plot = F,
                                 loadings_order=loadings_order,
                                 remove_l0f0 = F)
  plot2 = structure_plot_general(res$output[[i]]$fitted_model$ebpmf$fit_flash$L.pm,
                                 res$output[[i]]$fitted_model$ebpmf$fit_flash$F.pm,
                                 grouping =cell_names,
                                 title='ebpmf',
                                 print_plot = F,
                                 loadings_order=loadings_order)
  grid.arrange(plot1, plot0,plot2, nrow=3)
}
```


```{r}
fit = readRDS('/project2/mstephens/dongyue/poisson_mf/pbmc_fasttopics/init_tol_effect/nonnegLF_pe_inittol1e6_iter60.rds')
for(i in 1:4){
  plot(fit$sigma2,res$output[[i]]$fitted_model$ebpmf$sigma2,col='grey80',pch=20,xlab='true',ylab='estimated',main='variance estimation')
  abline(a=0,b=1)
}

```


## Large true variance, small init variance


```{r}
res = readRDS('/project2/mstephens/dongyue/poisson_mf/pbmc3k_simulation/simu_pbmc_fasttopics_large_true_var_small_init_var.rds')
K_hat = c()
for(i in 1:length(res$output)){
  K_hat = rbind(K_hat,c(res$output[[i]]$fitted_model$flash$n.factors,res$output[[i]]$fitted_model$ebpmf$fit_flash$n.factors))
}
colnames(K_hat) = c('flash','ebpmf')
K_hat
```

```{r}
loadings_order = get_loadings_order(res$sim_data$Loading,res$sim_data$Factor,grouping = cell_names,n_samples = 5000)
plot0=structure_plot_general(res$sim_data$Loading,res$sim_data$Factor,grouping =cell_names,title = 'True',print_plot = F,loadings_order = loadings_order) 
for(i in 1:length(res$output)){
  plot1 = structure_plot_general(res$output[[i]]$fitted_model$flash$L.pm,
                                 res$output[[i]]$fitted_model$flash$F.pm
                                 ,grouping=cell_names,
                                 title='flash',
                                 print_plot = F,
                                 loadings_order=loadings_order,
                                 remove_l0f0 = F)
  plot2 = structure_plot_general(res$output[[i]]$fitted_model$ebpmf$fit_flash$L.pm,
                                 res$output[[i]]$fitted_model$ebpmf$fit_flash$F.pm,
                                 grouping =cell_names,
                                 title='ebpmf',
                                 print_plot = F,
                                 loadings_order=loadings_order)
  grid.arrange(plot1, plot0,plot2, nrow=3)
}
```


```{r}
fit = readRDS('/project2/mstephens/dongyue/poisson_mf/pbmc_fasttopics/init_tol_effect/nonnegLF_pe_inittol1e2_iter60.rds')
for(i in 1:length(res$output)){
  plot(fit$sigma2,res$output[[i]]$fitted_model$ebpmf$sigma2,col='grey80',pch=20,xlab='true',ylab='estimated',main='variance estimation')
  abline(a=0,b=1)
}

```

## Large true variance, large init variance


```{r}
res = readRDS('/project2/mstephens/dongyue/poisson_mf/pbmc3k_simulation/simu_pbmc_fasttopics_large_true_var_large_init_var.rds')
K_hat = c()
for(i in c(1,3,4)){
  K_hat = rbind(K_hat,c(res$output[[i]]$fitted_model$flash$n.factors,res$output[[i]]$fitted_model$ebpmf$fit_flash$n.factors))
}
colnames(K_hat) = c('flash','ebpmf')
K_hat
```

```{r}
loadings_order = get_loadings_order(res$sim_data$Loading,res$sim_data$Factor,grouping = cell_names,n_samples = 5000)
plot0=structure_plot_general(res$sim_data$Loading,res$sim_data$Factor,grouping =cell_names,title = 'True',print_plot = F,loadings_order = loadings_order) 
for(i in c(1,3,4)){
  plot1 = structure_plot_general(res$output[[i]]$fitted_model$flash$L.pm,
                                 res$output[[i]]$fitted_model$flash$F.pm
                                 ,grouping=cell_names,
                                 title='flash',
                                 print_plot = F,
                                 loadings_order=loadings_order,
                                 remove_l0f0 = F)
  plot2 = structure_plot_general(res$output[[i]]$fitted_model$ebpmf$fit_flash$L.pm,
                                 res$output[[i]]$fitted_model$ebpmf$fit_flash$F.pm,
                                 grouping =cell_names,
                                 title='ebpmf',
                                 print_plot = F,
                                 loadings_order=loadings_order)
  grid.arrange(plot1, plot0,plot2, nrow=3)
}
```


```{r}
fit = readRDS('/project2/mstephens/dongyue/poisson_mf/pbmc_fasttopics/init_tol_effect/nonnegLF_pe_inittol1e2_iter60.rds')
for(i in c(1,3,4,5)){
  plot(fit$sigma2,res$output[[i]]$fitted_model$ebpmf$sigma2,col='grey80',pch=20,xlab='true',ylab='estimated',main='variance estimation')
  abline(a=0,b=1)
}

```

