---
title: "Simulation data from fitted model on pbmc data in fasttopics, nonnegtive L and F, true sigma2 = 0"
author: "DongyueXie"
date: "2023-02-26"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

So in a [previous simulation results](simu_study_of_init_variance.html), the true variances are those learnt from data. 

Now I want to see what if true variances are 0? I used the fitted model with relative large variance(init tol 1e-2).

The variances refers to $var(\epsilon_{ij})$ in 

\begin{equation}
        \begin{split}
            &y_{ij}\sim Poisson(s_i\lambda_{ij}),
            \\
            &\log\lambda_{ij} = f_{j0} + \sum_k l_{ik}f_{jk} + \epsilon_{ij},
            \\
            &l_{ik}\sim g_{k} ^{(l)},k=1,2,...,K,
            \\
            &f_{jk}\sim g_{k} ^{(f)},k=1,2,...,K.
        \end{split}
    \end{equation}

```{r}
library(fastTopics)
library(Matrix)
library(gridExtra)
source('code/poisson_STM/plot_factors.R')
source('code/poisson_STM/plot_factors_general.R')
source('code/poisson_STM/structure_plot.R')
source('code/poisson_STM/get_loadings_order.R')
data("pbmc_facs")
cell_names = pbmc_facs$samples$subpop
```

## small init variance

```{r}
res = readRDS('/project2/mstephens/dongyue/poisson_mf/pbmc3k_simulation/simu_pbmc_fasttopics_0_var_small_init_var.rds')
K_hat = c()
for(i in c(1,2,3,5)){
  K_hat = rbind(K_hat,c(res$output[[i]]$fitted_model$flash$n.factors,res$output[[i]]$fitted_model$ebpmf$fit_flash$n.factors,ncol(res$sim_data$Factor)))
}
colnames(K_hat) = c('flash','ebpmf','true K')
K_hat
```


```{r,warning=T}
loadings_order = get_loadings_order(res$sim_data$Loading,res$sim_data$Factor,grouping = cell_names,n_samples = 5000)
plot0=structure_plot_general(res$sim_data$Loading,res$sim_data$Factor,grouping =cell_names,title = 'True',print_plot = F,loadings_order=loadings_order) 
for(i in c(1,2,3,5)){
  plot1 = structure_plot_general(res$output[[i]]$fitted_model$flash$L.pm,
                                 res$output[[i]]$fitted_model$flash$F.pm
                                 ,grouping=cell_names,
                                 title='flash',
                                 print_plot = F,
                                 loadings_order=loadings_order,
                                 remove_l0f0 = F)
  plot2 = structure_plot_general(res$output[[i]]$fitted_model$ebpmf$fit_flash$L.pm,
                                 res$output[[i]]$fitted_model$ebpmf$fit_flash$F.pm,
                                 grouping =cell_names,
                                 title='ebpmf',
                                 print_plot = F,
                                 loadings_order=loadings_order)
  grid.arrange(plot1, plot0,plot2, nrow=3)
}
```


```{r}
for(i in c(1,2,3,5)){
  hist(res$output[[i]]$fitted_model$ebpmf$sigma2,breaks = 100,main='',xlab='var')
}
```

```{r}
for(i in c(1,2,3,5)){
  print(res$output[[i]]$fitted_model$ebpmf$run_time)
}
```

## large init variance

I used unimodal-nonnegative prior for this simulation because the point-exponential always ran into errors.

```{r}
res = readRDS('/project2/mstephens/dongyue/poisson_mf/pbmc3k_simulation/simu_pbmc_fasttopics_0_var_large_init_var.rds')
K_hat = c()
for(i in 1:5){
  K_hat = rbind(K_hat,c(res$output[[i]]$fitted_model$flash$n.factors,res$output[[i]]$fitted_model$ebpmf$fit_flash$n.factors,ncol(res$sim_data$Factor)))
}
colnames(K_hat) = c('flash','ebpmf','true K')
K_hat
```

```{r,warning=T}
loadings_order = get_loadings_order(res$sim_data$Loading,res$sim_data$Factor,grouping = cell_names,n_samples = 5000)
plot0=structure_plot_general(res$sim_data$Loading,res$sim_data$Factor,grouping =cell_names,title = 'True',print_plot = F,loadings_order=loadings_order) 
for(i in 1:5){
  plot1 = structure_plot_general(res$output[[i]]$fitted_model$flash$L.pm,
                                 res$output[[i]]$fitted_model$flash$F.pm
                                 ,grouping=cell_names,
                                 title='flash',
                                 print_plot = F,
                                 loadings_order=loadings_order,
                                 remove_l0f0 = F)
  plot2 = structure_plot_general(res$output[[i]]$fitted_model$ebpmf$fit_flash$L.pm,
                                 res$output[[i]]$fitted_model$ebpmf$fit_flash$F.pm,
                                 grouping =cell_names,
                                 title='ebpmf',
                                 print_plot = F,
                                 loadings_order=loadings_order)
  grid.arrange(plot1, plot0,plot2, nrow=3)
}
```


```{r}
for(i in 1:5){
  hist(res$output[[i]]$fitted_model$ebpmf$sigma2,breaks = 100,main='',xlab='var')
}
```

```{r}
for(i in 1:5){
  print(res$output[[i]]$fitted_model$ebpmf$run_time)
}
```
