---
title: "Smoothing Poisson spikefunction benchmark"
author: "DongyueXie"
date: "2022-11-18"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

**I picked up some methods that are performing better and reduce the plot size for better illustration. See [here](smooth_spike_benchmark.html) for full version**

```{r}
rm_methods = c('lik_exp_smashpoi_nug_top','smash_two_step_homo','lik_exp_smashpoi_nug','lik_exp_smashpoi_nug_top','vst_nug','vst_nug_top','split_runmed_dwt','split_true','split_log1px_ndwt_fix_nug','split_log1px_ndwt','split_runmed_ndwt_fix_nug','lik_exp_mle_smooth','lik_exp_mle_nug_top','lik_exp_mle_nug','split_runmed_ndwt_top')
include_methods = c('vst_smooth','lik_exp_smashpoi_smooth','split_runmed_ndwt','smash_two_step_hetero','smash')
```

In this analysis we look at the results from generating data as 

$$y_i\sim Poisson(\exp(\mu_i+\epsilon_i)),\epsilon_i\sim N(0,\sigma^2).$$
The $\mu$ is smooth and has block structures as follows:

```{r}
library(tidyverse)
out = readRDS('output/poisson_smooth_simulation/poisson_smooth/spike30_n_1024_snr_1_count_size_3.rds')
plot(out$sim_data$b,type='l',ylab='')
```

We set $n=1024$, repeat 30 times, and count size $=max(\exp(b))\in \{3,10,100\}$, and $snr = var(b)/\sigma^2\in \{1,3\}$.


```{r}
source('code/poisson_smooth/simulation_summary_smooth.R')
```


## count size = 3

### snr = 1

```{r}
out = readRDS('output/poisson_smooth_simulation/poisson_smooth/spike30_n_1024_snr_1_count_size_3.rds')
```

```{r}
get_mse_mean(out,rm_method = rm_methods)
get_runtime(out,rm_method = rm_methods)
```

```{r}
for(method in include_methods){
  plot_all_curves(out,method=method)
}

```


### snr = 3

```{r}
out = readRDS('output/poisson_smooth_simulation/poisson_smooth/spike30_n_1024_snr_3_count_size_3.rds')
```

```{r}
get_mse_mean(out,rm_method = rm_methods)
get_runtime(out,rm_method = rm_methods)
```

```{r}
for(method in include_methods){
  plot_all_curves(out,method=method)
}

```


## count size = 10

### snr = 1

```{r}
out = readRDS('output/poisson_smooth_simulation/poisson_smooth/spike30_n_1024_snr_1_count_size_10.rds')
```

```{r}
get_mse_mean(out,rm_method = rm_methods)
get_runtime(out,rm_method = rm_methods)
```

```{r}
for(method in include_methods){
  plot_all_curves(out,method=method)
}

```


### snr = 3

```{r}
out = readRDS('output/poisson_smooth_simulation/poisson_smooth/spike30_n_1024_snr_3_count_size_10.rds')
```
```{r}
get_mse_mean(out,rm_method = rm_methods)
get_runtime(out,rm_method = rm_methods)
```

```{r}
for(method in include_methods){
  plot_all_curves(out,method=method)
}

```


## count size = 100

### snr = 1

```{r}
out = readRDS('output/poisson_smooth_simulation/poisson_smooth/spike30_n_1024_snr_1_count_size_100.rds')
```

```{r}
get_mse_mean(out,rm_method = rm_methods)
get_runtime(out,rm_method = rm_methods)
```

```{r}
for(method in include_methods){
  plot_all_curves(out,method=method)
}

```

### snr = 3

```{r}
out = readRDS('output/poisson_smooth_simulation/poisson_smooth/spike30_n_1024_snr_3_count_size_100.rds')
```

```{r}
get_mse_mean(out,rm_method = rm_methods)
get_runtime(out,rm_method = rm_methods)
```

```{r}
for(method in include_methods){
  plot_all_curves(out,method=method)
}

```
