---
title: "splitting method in low count smoothing"
author: "DongyueXie"
date: "2022-11-21"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Update: Some observations. 

When fixing $\sigma^2$ at truth, dwt can find the highest peak and second highest peak, when initializing at true $\mu$. But ndwt always over-smooth the curve to constant. Maybe ndwt has a stronger smooth effect than dwt? One resaon is that the ndwt curve is always below the points so the points are always shunk towards smaller values.





It seems that the splitting method may over-smooth the low count sequence.

We load the results from Spike function with small count size.

```{r}
library(smashrgen)
out = readRDS('output/poisson_smooth_simulation/poisson_smooth/spike30_n_1024_snr_1_count_size_3.rds')
```

We start with fixing the nugget effect at truth.

We initialize the $\bar\mu$ at the true $\mu$.

```{r,warning=F}
i = 4
x = out$sim_data$X[i,]
fit = pois_smooth_split(x,wave_trans = 'dwt',sigma2_init = out$sim_data$sigma2,est_sigma2 = F,Emu_init = log(out$sim_data$latentX[i,]),maxiter = 10,plot_updates = T)
plot(x,col='grey80')
lines(exp(out$sim_data$b),col='grey80')
lines(fit$posterior$mean_smooth)
```

```{r,warning=F}
i = 4
x = out$sim_data$X[i,]
fit = pois_smooth_split(x,wave_trans = 'ndwt',sigma2_init = out$sim_data$sigma2,est_sigma2 = F,Emu_init = log(out$sim_data$latentX[i,]),maxiter = 10,plot_updates = T)
plot(x,col='grey80')
lines(exp(out$sim_data$b),col='grey80')
lines(fit$posterior$mean_smooth)
```






