---
title: "splitting PMF 1 backfitting in init"
author: "DongyueXie"
date: "2022-12-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

The results here show that using 1 back-fitting in flash initialization does not affect the final results.

```{r}
rmse = function(x,y){return(sqrt(mean((x-y)^2)))}
res = readRDS('/project2/mstephens/dongyue/poisson_mf/PMF10_K3simu_pbmc_3cells_1backinit.rds')
```

We first look at the number factors recovered from both methods. The true $K$ is 3.

```{r}
K_hat = c()
for(i in 1:length(res$output)){
  K_hat = rbind(K_hat,c(res$output[[i]]$fitted_model$flash$n.factors,res$output[[i]]$fitted_model$splitting$fit_flash$n.factors))
}
colnames(K_hat) = c('flash','splittingPMF')
K_hat
```

Next we compare $\hat L\hat F'$ and true $LF'$.

```{r}
fit = readRDS('/project2/mstephens/dongyue/poisson_mf/pbmc_3cells_Sij.rds')
kset = order(fit$fit$fit_flash$pve,decreasing = TRUE)[1:3]
Ltrue = fit$fit$fit_flash$L.pm[,kset]
Ftrue = fit$fit$fit_flash$F.pm[,kset]
Mu_true = tcrossprod(Ltrue,Ftrue)
```

```{r}
rmses= c()
for(i in 1:length(res$output)){
  rmses = rbind(rmses,c(rmse(Mu_true,fitted(res$output[[i]]$fitted_model$flash)),rmse(Mu_true,fitted(res$output[[i]]$fitted_model$splitting$fit_flash))))
}
colnames(rmses) = c('flash','splittingPMF')
rmses
```

```{r,fig.width=10,fig.height=5,eval=F}
par(mfrow=c(2,1))
for(i in 1:length(res$output)){
  plot(fitted(res$output[[i]]$fitted_model$flash),Mu_true,col='grey80',xlab='fitted',ylab='LF',main='flash')
  abline(a=0,b=1)
  plot(fitted(res$output[[i]]$fitted_model$splitting$fit_flash),Mu_true,col='grey80',xlab='fitted',ylab='LF',main='splitting')
  abline(a=0,b=1)
}
par(mfrow=c(1,1))
```

Next we look at how the structures of L and F are recovered by both methods.

We first plot loadings.

```{r}
library(fastTopics)
library(Matrix)
library(stm)
require(gridExtra)
data(pbmc_facs)
counts <- pbmc_facs$counts
table(pbmc_facs$samples$subpop)
## use only B cell and NK cell and CD34+
cells = pbmc_facs$samples$subpop%in%c('B cell', 'NK cell','CD34+')
counts = counts[cells,]
# filter out genes that has few expressions(3% cells)
genes = (colSums(counts>0) > 0.03*dim(counts)[1])
cell_names = pbmc_facs$samples$subpop[cells]
source('code/poisson_STM/plot_factors.R')
```

```{r,fig.width=16,fig.height=4}
plot0=plot.factors(fit$fit$fit_flash,cell.types=cell_names,kset=kset,title='True Loadings')
for(i in 1:length(res$output)){
  plot1 = plot.factors(res$output[[i]]$fitted_model$flash,cell.types=cell_names,title='flash')
  plot2 = plot.factors(res$output[[i]]$fitted_model$splitting$fit_flash,cell.types=cell_names,title='splittingPMF')
  grid.arrange(plot1, plot0,plot2, ncol=3)
}

```

## misc

number of iterations until convergence

```{r}
for(i in 1:length(res$output)){
   print(length(res$output[[i]]$fitted_model$splitting$eblo_trace))
}
```


run time analysis

```{r}
for(i in 1:length(res$output)){
   print((res$output[[i]]$fitted_model$splitting$run_time))
}
```

run time break down

```{r}
tt = c()
for(i in 1:length(res$output)){
   tt = rbind(tt,unlist(lapply(res$output[[i]]$fitted_model$splitting$run_time_break_down,mean)))
}
apply(tt,2,mean)
```

Plot K trace

```{r}
plot(res$output[[i]]$fitted_model$splitting$K_trace,ylab='K',xlab='iterations',type='l',col='grey80')
for(i in 1:length(res$output)){
   lines(res$output[[i]]$fitted_model$splitting$K_trace,col=i)
}
```


Maybe omit greedy step after the K is stabilized? to reduce computation time.
