---
title: "t prior benchmark revised"
author: "DongyueXie"
date: "2022-11-10"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r}
library(vebpm)
library(ggplot2)
library(tidyverse)
source('code/poisson_mean/simulation_summary.R')
```

Generated $n=1000$ observations from \[y_i\sim Poisson(\lambda_i),\lambda_i\sim g(\cdot).\]

Look at MSE of $\hat\lambda$ and estimate of $\log\lambda$.

See [here](t_prior_benchmark.html) for full results. 

Note that nb_lb and nb_pg are equivalent and I set r=1000 for the former and r=100 for the latter for comparisons.

## $\lambda_j \sim \pi_0\delta_1 + \pi_1(2+|t_3|)$

```{r}
out = readRDS('output/poisson_mean_simulation/poisson_mean/t_prior31_n_1000_df_3_delta_1.rds')
```

```{r}
get_summary_mean(out,rm_method = c('ash_pois_log',"ebpm_exp_mixture",'GMGM_pointmass'))
get_summary_mean_log(out,rm_method = c('ash_pois_log',"ebpm_exp_mixture",'GMGM_pointmass'))
```

## $\lambda_j \sim \pi_0\delta_2 + \pi_1(4+|t_3|)$

```{r}
out = readRDS('output/poisson_mean_simulation/poisson_mean/t_prior31_n_1000_df_3_delta_2.rds')
```

```{r}
get_summary_mean(out,rm_method = c('ash_pois_log',"ebpm_exp_mixture",'GMGM_pointmass'))
get_summary_mean_log(out,rm_method = c('ash_pois_log',"ebpm_exp_mixture",'GMGM_pointmass'))
```

## $\lambda_j \sim \pi_0\delta_5 + \pi_1(10+|t_3|)$

```{r}
out = readRDS('output/poisson_mean_simulation/poisson_mean/t_prior31_n_1000_df_3_delta_5.rds')
```

```{r}
get_summary_mean(out,rm_method = c('ash_pois_log',"ebpm_exp_mixture",'GMGM_pointmass'))
get_summary_mean_log(out,rm_method = c('ash_pois_log',"ebpm_exp_mixture",'GMGM_pointmass'))
```

```{r}
  mse_all = c()
  for(i in 1:length(out$output)){
    mse_all = rbind(mse_all,out$output[[i]]$MSE_mean)
  }
```

```{r}
mse_all[,10]
```

```{r}
x = out$sim_data$X[1,]
plot(x,col='grey80')
lines(out$output[[1]]$fitted_model$penalty_inversion$posterior$mean)
```


```{r}
which(x==0)
which(out$output[[1]]$fitted_model$penalty_inversion$posterior$mean>=30)
```

So those extreme large estimates correspond to $x=0$.







